---
title: "Practical: Decision Trees, Random Forests, and XGBoost"
subtitle: "Tree-based classification on the Wisconsin Diagnostic Breast Cancer dataset"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
---


# Setup

```{r}
library(tidyverse)
library(broom)
library(yardstick)
library(pROC)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
```



# pandas equivalent:
# from sklearn.tree import DecisionTreeClassifier, plot_tree
# from sklearn.ensemble import RandomForestClassifier
# from xgboost import XGBClassifier

---

# 1. Why decision trees?

Logistic regression is powerful, but it assumes a relatively simple relationship between predictors and log-odds.

Decision trees work by asking a sequence of simple questions:

```text
Is cell_size > 3?
  yes -> Is bare_nuclei > 2?
    yes -> malignant
    no  -> benign
  no -> benign
```

Advantages:

- easy to explain;
- can capture nonlinear relationships;
- can capture interactions;
- no need to scale variables;
- works with numeric and categorical variables;
- useful baseline model.

Disadvantages:

- unstable;
- prone to overfitting;
- usually less accurate than ensembles;
- small data changes can produce different trees.

---

# 2. Dataset: Wisconsin Diagnostic Breast Cancer

We will use the same WDBC dataset from the PCA practical.

Each row describes a breast mass sample.

Target:

- `M`: malignant;
- `B`: benign.

Predictors:

- 30 quantitative features extracted from digitized cell nucleus images.

## Load data

```{r}
wdbc_raw <- read_csv("Class 4/breast-cancer.csv")

glimpse(wdbc_raw)

names(wdbc_raw) <- names(wdbc_raw) %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("-", "_") %>%
  str_replace_all("\\.", "_") %>%
  str_to_lower()

wdbc <- wdbc_raw %>%
  mutate(
    diagnosis = factor(diagnosis, levels = c("M", "B"))
  ) %>%
  select(-id)

glimpse(wdbc)
```


## Class balance

```{r}
wdbc %>%
  count(diagnosis) %>%
  mutate(percent = n / sum(n) * 100)
```

```{r}
ggplot(wdbc, aes(x = diagnosis, fill = diagnosis)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Class balance",
    x = "Diagnosis",
    y = "Count"
  )
```

---

# 3. Train/test split

We keep the same principle as before:

- train set: used to fit the model;
- test set: used only for final evaluation.

This helps estimate performance on unseen future data.

```{r}
train_data <- wdbc %>%
  group_by(diagnosis) %>%
  slice_sample(prop = 0.75) %>%
  ungroup()

test_data <- wdbc %>%
  anti_join(train_data, by = names(wdbc))

train_data %>% count(diagnosis)
test_data %>% count(diagnosis)
```

---

# 4. Decision tree intuition

Decision trees split data into smaller and smaller groups.

At each step, the algorithm chooses:

- a variable;
- a threshold;
- a split that best separates the classes.

Example split:

```text
concave_points_worst < 0.14
```

The algorithm asks:

> Does this split create purer groups?

A pure group contains mostly one class.

## Gini impurity

One common splitting criterion is **Gini impurity**.

For a binary classification node:

$$
Gini = 1 - p_1^2 - p_2^2
$$

Where:

- `p1` is the proportion of class 1;
- `p2` is the proportion of class 2.

Interpretation:

- Gini = 0 means perfectly pure;
- higher Gini means more mixed.

## Visual Gini example

```{r}
gini_data <- tibble(
  p_malignant = seq(0, 1, length.out = 200),
  p_benign = 1 - p_malignant,
  gini = 1 - p_malignant^2 - p_benign^2
)

ggplot(gini_data, aes(x = p_malignant, y = gini)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Gini impurity for binary classification",
    x = "Proportion malignant",
    y = "Gini impurity"
  )
```

Question:

> At which class proportion is the node most mixed?


---

# 5. Fit a basic decision tree

We use `rpart()`.

```{r}
tree_model <- rpart(
  diagnosis ~ .,
  data = train_data,
  method = "class"
)

tree_model
```

# pandas equivalent:
# tree = DecisionTreeClassifier(random_state=42)
# tree.fit(X_train, y_train)

---

## Plot the tree

```{r}
rpart.plot(
  tree_model,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  main = "Decision tree for WDBC classification"
)
```

How to read the tree:

- each internal node is a decision rule;
- each branch is an answer;
- each leaf is a final prediction;
- numbers usually show class probability and percentage of samples.

---

## Feature importance from tree

```{r}
tree_model$variable.importance
```

```{r}
tree_importance <- tibble(
  feature = names(tree_model$variable.importance),
  importance = as.numeric(tree_model$variable.importance)
)

ggplot(tree_importance, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Decision tree variable importance",
    x = "Feature",
    y = "Importance"
  )
```

---

# 6. Evaluate decision tree

```{r}
tree_predictions_class <- predict(
  tree_model,
  newdata = test_data,
  type = "class"
)

tree_predictions_prob <- predict(
  tree_model,
  newdata = test_data,
  type = "prob"
)[, "M"]

tree_test_predictions <- test_data %>%
  mutate(
    predicted_class = factor(tree_predictions_class, levels = c("M", "B")),
    prob_malignant = tree_predictions_prob
  )

tree_test_predictions %>%
  select(diagnosis, predicted_class, prob_malignant) %>%
  head()
```

---

## Confusion matrix

```{r}
table(
  Truth = tree_test_predictions$diagnosis,
  Prediction = tree_test_predictions$predicted_class
)
```

---

## Classification metrics

```{r}
tree_metric_data <- tree_test_predictions %>%
  mutate(
    diagnosis = factor(diagnosis, levels = c("M", "B")),
    predicted_class = factor(predicted_class, levels = c("M", "B"))
  )

metric_set(accuracy, precision, recall, spec, f_meas)(
  tree_metric_data,
  truth = diagnosis,
  estimate = predicted_class
)
```

---

## ROC AUC

```{r}
tree_auc <- roc(
  response = tree_test_predictions$diagnosis,
  predictor = tree_test_predictions$prob_malignant,
  levels = c("B", "M")
) %>%
  auc()

tree_auc
```

---

# 7. Overfitting in decision trees

Decision trees can easily become too complex.

A very deep tree may memorize training data instead of learning general rules.

Symptoms:

- training performance is very high;
- test performance is worse;
- tree has too many small leaves;
- rules become too specific.

---

## Fit a larger tree

```{r}
large_tree <- rpart(
  diagnosis ~ .,
  data = train_data,
  method = "class",
  control = rpart.control(
    cp = 0.0001,
    minsplit = 2,
    maxdepth = 10
  )
)

rpart.plot(
  large_tree,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  main = "Large decision tree"
)
```

---

## Compare tree complexity

```{r}
printcp(large_tree)
```

The `cp` table shows how error changes as the tree is pruned.

---

## Plot complexity parameter

```{r}
plotcp(large_tree)
```

The goal is not always to use the largest tree.

We often prefer a smaller tree with similar validation error.

---

# 8. Pruning a decision tree

Pruning removes branches that do not improve performance enough.

A common approach:

1. Fit a large tree.
2. Choose a complexity parameter `cp`.
3. Prune the tree.
4. Evaluate the pruned tree.

---

## Choose best CP

```{r}
best_cp <- large_tree$cptable %>%
  as.data.frame() %>%
  arrange(xerror) %>%
  dplyr::slice(1) %>%
  pull(CP)

best_cp
```

---

## Prune the tree

```{r}
pruned_tree <- prune(
  large_tree,
  cp = best_cp
)

rpart.plot(
  pruned_tree,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  main = "Pruned decision tree"
)
```

---

## Evaluate pruned tree

```{r}
pruned_predictions_class <- predict(
  pruned_tree,
  newdata = test_data,
  type = "class"
)

pruned_predictions_prob <- predict(
  pruned_tree,
  newdata = test_data,
  type = "prob"
)[, "M"]

pruned_test_predictions <- test_data %>%
  mutate(
    predicted_class = factor(pruned_predictions_class, levels = c("M", "B")),
    prob_malignant = pruned_predictions_prob
  )

pruned_metric_data <- pruned_test_predictions %>%
  mutate(
    diagnosis = factor(diagnosis, levels = c("M", "B")),
    predicted_class = factor(predicted_class, levels = c("M", "B"))
  )

metric_set(accuracy, precision, recall, spec, f_meas)(
  pruned_metric_data,
  truth = diagnosis,
  estimate = predicted_class
)

pruned_auc <- roc(
  response = pruned_test_predictions$diagnosis,
  predictor = pruned_test_predictions$prob_malignant,
  levels = c("B", "M")
) %>%
  auc()

pruned_auc
```

---

# 9. Compare decision tree models

```{r}
evaluate_binary_model <- function(data, truth_col, class_col, prob_col) {
  eval_data <- tibble(
    truth = factor(data[[truth_col]], levels = c("M", "B")),
    estimate = factor(data[[class_col]], levels = c("M", "B")),
    prob = data[[prob_col]]
  )

  metrics <- metric_set(accuracy, precision, recall, spec, f_meas)(
    eval_data,
    truth = truth,
    estimate = estimate
  )

  auc_value <- roc(
    response = eval_data$truth,
    predictor = eval_data$prob,
    levels = c("B", "M")
  ) %>%
    auc() %>%
    as.numeric()

  metrics %>%
    select(.metric, .estimate) %>%
    pivot_wider(names_from = .metric, values_from = .estimate) %>%
    mutate(auc = auc_value)
}

tree_results <- bind_rows(
  evaluate_binary_model(tree_test_predictions, "diagnosis", "predicted_class", "prob_malignant") %>%
    mutate(model = "Basic tree"),
  evaluate_binary_model(pruned_test_predictions, "diagnosis", "predicted_class", "prob_malignant") %>%
    mutate(model = "Pruned tree")
) %>%
  select(model, accuracy, precision, recall, spec, f_meas, auc)

tree_results
```


---

# 10. Random Forest

A random forest is an ensemble of many decision trees.

Instead of relying on one tree, it builds many trees and averages their predictions.

Core ideas:

1. Bootstrap sampling: each tree is trained on a random sample of rows.
2. Random feature selection: each split considers only a random subset of variables.
3. Voting/averaging: final prediction combines many trees.

Why it helps:

- reduces overfitting;
- improves stability;
- usually gives better performance than a single tree;
- handles nonlinearities and interactions.

Main downside:

- less interpretable than one tree;
- more computationally expensive;
- predictions are harder to explain.

---

## Fit random forest

```{r}
set.seed(2026)

rf_model <- randomForest(
  diagnosis ~ .,
  data = train_data,
  ntree = 500,
  importance = TRUE
)

rf_model
```

# pandas equivalent:
# rf = RandomForestClassifier(n_estimators=500, random_state=42)
# rf.fit(X_train, y_train)

---

## Random forest predictions

```{r}
rf_predictions_class <- predict(
  rf_model,
  newdata = test_data,
  type = "response"
)

rf_predictions_prob <- predict(
  rf_model,
  newdata = test_data,
  type = "prob"
)[, "M"]

rf_test_predictions <- test_data %>%
  mutate(
    predicted_class = factor(rf_predictions_class, levels = c("M", "B")),
    prob_malignant = rf_predictions_prob
  )
```

---

## Evaluate random forest

```{r}
rf_results <- evaluate_binary_model(
  rf_test_predictions,
  "diagnosis",
  "predicted_class",
  "prob_malignant"
) %>%
  mutate(model = "Random forest") %>%
  select(model, accuracy, precision, recall, spec, f_meas, auc)

rf_results
```

---

## Random forest feature importance

```{r}
importance(rf_model)
```

```{r}
rf_importance <- importance(rf_model) %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  as_tibble()

rf_importance %>%
  arrange(desc(MeanDecreaseGini)) %>%
  slice(1:15) %>%
  ggplot(aes(x = reorder(feature, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Random forest feature importance",
    x = "Feature",
    y = "Mean decrease in Gini"
  )
```

---

# 11. XGBoost: basic idea

XGBoost is a gradient boosting method.

It also builds many trees, but differently from random forest.

Random forest:

- builds many trees independently;
- averages them;
- reduces variance.

Gradient boosting / XGBoost:

- builds trees sequentially;
- each new tree tries to fix previous model errors;
- often achieves very strong predictive performance.

Simplified idea:

```text
Model 1 makes errors
Model 2 learns from errors of Model 1
Model 3 learns from errors of Model 1 + Model 2
...
Final model = many weak trees combined
```

Advantages:

- very strong tabular-data performance;
- handles nonlinear relationships;
- handles interactions;
- flexible regularization;
- often wins machine learning competitions.

Disadvantages:

- many hyperparameters;
- more difficult to interpret;
- can overfit if tuned poorly;
- requires careful validation.

---

# 12. Optional: Fit XGBoost

Run this section only if `xgboost` is installed.

XGBoost requires numeric matrix input and numeric labels.

```{r}
library(xgboost)

x_train <- train_data %>% select(-diagnosis) %>% as.matrix()

 x_test <- test_data %>%select(-diagnosis) %>%as.matrix()

y_train <- if_else(train_data$diagnosis == "M", 1, 0)
y_test <- if_else(test_data$diagnosis == "M", 1, 0)

dtrain <- xgb.DMatrix(data = x_train, label = y_train)
dtest <- xgb.DMatrix(data = x_test, label = y_test)

xgb_params <- list(objective = "binary:logistic", eval_metric = "auc", max_depth = 3, eta = 0.1, subsample = 0.8, colsample_bytree = 0.8)


xgb_model <- xgb.train(params = xgb_params, data = dtrain, nrounds = 100, verbose = 0)

xgb_prob <- predict(xgb_model, dtest)

xgb_test_predictions <- test_data %>%
   mutate(
     prob_malignant = xgb_prob,
     predicted_class = factor(
       if_else(prob_malignant >= 0.5, "M", "B"),
       levels = c("M", "B")))
xgb_results <- evaluate_binary_model(
   xgb_test_predictions,
   "diagnosis",
   "predicted_class",
   "prob_malignant") %>%
   mutate(model = "XGBoost") %>%
   select(model, accuracy, precision, recall, spec, f_meas, auc)

xgb_results
```

---

## Optional: XGBoost feature importance

```{r}
xgb_importance <- xgb.importance(
   feature_names = colnames(x_train),
   model = xgb_model)

xgb_importance %>%
   as_tibble() %>%
   dplyr::slice(1:15) %>%
   ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
   geom_col(fill = "purple") +
   coord_flip() +
   theme_minimal() + labs(title = "XGBoost feature importance", x = "Feature", y = "Gain")
```

---

# 13. Compare all models

If you ran only tree and random forest:

```{r}
bind_rows(
  tree_results,
  rf_results,
  xgb_results
) %>%
  arrange(desc(auc))
```



## Model comparison conceptually

| Model | Strength | Weakness |
|---|---|---|
| Decision tree | Interpretable rules | Overfits easily |
| Random forest | Strong and stable | Less interpretable |
| XGBoost | Very strong performance | More tuning required |
| Logistic regression | Simple and interpretable | Mostly linear decision boundary |

---

# 14. Threshold discussion

For medical screening, the threshold is not only a technical choice.

If false negatives are very dangerous, we may prefer:

- lower threshold;
- higher recall;
- more false positives.

If false positives are very costly, we may prefer:

- higher threshold;
- higher precision;
- fewer false positives.

Example for random forest:

```{r}
threshold_results <- tibble(
  threshold = seq(0.1, 0.9, by = 0.1)
) %>%
  mutate(
    predictions = map(
      threshold,
      ~ rf_test_predictions %>%
        mutate(
          predicted_class = factor(
            if_else(prob_malignant >= .x, "M", "B"),
            levels = c("M", "B")
          ),
          diagnosis = factor(diagnosis, levels = c("M", "B"))
        )
    ),
    accuracy = map_dbl(predictions, ~ accuracy(.x, truth = diagnosis, estimate = predicted_class)$.estimate),
    precision = map_dbl(predictions, ~ precision(.x, truth = diagnosis, estimate = predicted_class)$.estimate),
    recall = map_dbl(predictions, ~ recall(.x, truth = diagnosis, estimate = predicted_class)$.estimate),
    specificity = map_dbl(predictions, ~ spec(.x, truth = diagnosis, estimate = predicted_class)$.estimate),
    f1 = map_dbl(predictions, ~ f_meas(.x, truth = diagnosis, estimate = predicted_class)$.estimate)
  ) %>%
  select(-predictions)

threshold_results
```

```{r}
threshold_results %>%
  pivot_longer(
    cols = c(accuracy, precision, recall, specificity, f1),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = threshold, y = value, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Effect of threshold on random forest metrics",
    x = "Threshold",
    y = "Metric value",
    color = "Metric"
  )
```

---

# 15. challenge

Using the WDBC dataset:

1. Train a decision tree with different `cp` values.
2. Compare tree size and test AUC.
3. Train a random forest with:
   - `ntree = 100`;
   - `ntree = 500`.
4. Compare random forest feature importance with decision tree feature importance.
5. Decide which model you would use for:
   - explanation;
   - prediction.

```{r}

```

