---
title: "Practical: Gradient Descent and Logistic Regression"
subtitle: "Optimization, Breast Cancer Wisconsin classification, train/test split, and metrics"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
---


# Setup

```{r}
library(tidyverse)
library(broom)
library(yardstick)
library(pROC)
library(mlbench)
```

# pandas equivalent:
# import pandas as pd
# import numpy as np
# import seaborn as sns
# import matplotlib.pyplot as plt
# import statsmodels.api as sm
# from sklearn.model_selection import train_test_split
# from sklearn.datasets import load_breast_cancer
# from sklearn.linear_model import LogisticRegression
# from sklearn.metrics import confusion_matrix, accuracy_score, precision_score, recall_score, f1_score, roc_curve, roc_auc_score


# 1. Why do we need gradient descent?

Many machine learning models are trained by finding parameters that minimize a loss function.

In linear regression, we write:

$$
\hat{y} = \beta_0 + \beta_1x
$$

But how do we find the best values of `beta_0` and `beta_1`?

We need a **loss function**.

For regression, a common loss function is mean squared error:

$$
MSE = \frac{1}{n}\sum(y_i - \hat{y}_i)^2
$$

The model tries to find parameters that make this loss as small as possible.

---

## Visual idea of a loss function

Let us imagine that we have only one parameter, `theta`.

The loss function is:

$$
J(\theta) = (\theta - 2)^2 + 1
$$

The minimum is at `theta = 2`.

```{r}
loss_data <- tibble(
  theta = seq(-5, 8, length.out = 300),
  loss = (theta - 2)^2 + 1
)

ggplot(loss_data, aes(x = theta, y = loss)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 2, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "A simple loss function",
    subtitle = "The best parameter value is where the loss is minimal",
    x = "Parameter theta",
    y = "Loss"
  )
```

---

# 2. Gradient descent intuition

Gradient descent is an iterative optimization method.

The update rule is:

$$
\theta_{new} = \theta_{old} - \eta \nabla J(\theta)
$$

Where:

- `theta` is the parameter;
- `eta` is the learning rate;
- `gradient` is the slope of the loss function;
- each step moves the parameter toward lower loss.

For our toy loss function:

$$
J(\theta) = (\theta - 2)^2 + 1
$$

the gradient is:

$$
\nabla J(\theta) = 2(\theta - 2)
$$

---

## Manual gradient descent simulation

```{r}
theta <- -4
learning_rate <- 0.2
n_steps <- 15

gd_path <- tibble(
  step = 0,
  theta = theta,
  loss = (theta - 2)^2 + 1
)

for (i in 1:n_steps) {
  gradient <- 2 * (theta - 2)
  theta <- theta - learning_rate * gradient

  gd_path <- bind_rows(
    gd_path,
    tibble(
      step = i,
      theta = theta,
      loss = (theta - 2)^2 + 1
    )
  )
}

gd_path
```

```{r}
ggplot(loss_data, aes(x = theta, y = loss)) +
  geom_line(linewidth = 1.2) +
  geom_point(
    data = gd_path,
    aes(x = theta, y = loss, color = step),
    size = 3
  ) +
  geom_path(
    data = gd_path,
    aes(x = theta, y = loss),
    arrow = arrow(length = unit(0.15, "cm")),
    linewidth = 0.8
  ) +
  theme_minimal() +
  labs(
    title = "Gradient descent path",
    subtitle = "The parameter moves step by step toward the minimum",
    x = "Parameter theta",
    y = "Loss",
    color = "Step"
  )
```

---

## Learning rate effect

The learning rate controls step size.

If it is too small, optimization is slow.  
If it is too large, optimization can overshoot or diverge.

```{r}
simulate_gd <- function(start_theta, learning_rate, n_steps = 15) {
  theta <- start_theta

  out <- tibble(
    step = 0,
    theta = theta,
    loss = (theta - 2)^2 + 1,
    learning_rate = learning_rate
  )

  for (i in 1:n_steps) {
    gradient <- 2 * (theta - 2)
    theta <- theta - learning_rate * gradient

    out <- bind_rows(
      out,
      tibble(
        step = i,
        theta = theta,
        loss = (theta - 2)^2 + 1,
        learning_rate = learning_rate
      )
    )
  }

  out
}

gd_rates <- bind_rows(
  simulate_gd(-4, 0.05),
  simulate_gd(-4, 0.2),
  simulate_gd(-4, 0.9)
)

ggplot(gd_rates, aes(x = step, y = loss, color = as.factor(learning_rate))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Effect of learning rate",
    x = "Gradient descent step",
    y = "Loss",
    color = "Learning rate"
  )
```



# 3. From regression to classification

Linear regression predicts numeric values:

$$
\hat{y} = \beta_0 + \beta_1x
$$

But classification needs class labels.

For example:

- benign / malignant;
- disease / no disease;
- spam / not spam;
- responder / non-responder.

If we use linear regression for binary classes, it can predict invalid values like:

- `-0.2`
- `1.4`

Instead, we want a model that predicts probabilities between 0 and 1.

---

# 4. Sigmoid function

The sigmoid function converts any real number into a value between 0 and 1.

$$
\sigma(z) = \frac{1}{1 + e^{-z}}
$$

```{r}
sigmoid_data <- tibble(
  z = seq(-10, 10, length.out = 300),
  probability = 1 / (1 + exp(-z))
)

ggplot(sigmoid_data, aes(x = z, y = probability)) +
  geom_line(linewidth = 1.3, color = "steelblue") +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Sigmoid function",
    subtitle = "Any real value is transformed into a probability",
    x = "Linear predictor z",
    y = "Predicted probability"
  )
```

---

# 5. Logistic regression model

Logistic regression has two parts.

First, a linear predictor:

$$
z = \beta_0 + \beta_1x_1 + \beta_2x_2 + ...
$$

Then the sigmoid transformation:

$$
P(Y = 1) = \frac{1}{1 + e^{-z}}
$$

So logistic regression is still a linear model, but it predicts probability.

For binary classification, logistic regression is usually fitted by minimizing **log loss**, not MSE.

---

# 6. Real dataset: Breast Cancer Wisconsin

We will use the Breast Cancer Wisconsin dataset from the `mlbench` package.

Each row describes a cell sample.  
The target is whether the sample is:

- benign;
- malignant.

```{r}
data(BreastCancer)

bc_raw <- BreastCancer %>%
  as_tibble()

glimpse(bc_raw)
```

---

## Dataset cleaning

The original dataset has an ID column and some variables stored as factors.

We will:

1. remove the ID column;
2. convert predictors to numeric;
3. remove rows with missing values;
4. define the positive class as `malignant`.

```{r}
bc <- bc_raw %>%
  select(-Id) %>%
  mutate(
    Class = factor(Class, levels = c("malignant", "benign"))
  ) %>%
  mutate(
    across(
      .cols = -Class,
      .fns = ~ as.numeric(as.character(.x))
    )
  ) %>%
  drop_na()

glimpse(bc)
```

# pandas equivalent:
# data = load_breast_cancer()
# df = pd.DataFrame(data.data, columns=data.feature_names)
# df["diagnosis"] = data.target
# df["diagnosis"] = np.where(df["diagnosis"] == 0, "malignant", "benign")

---

## Target class balance

```{r}
bc %>%
  count(Class) %>%
  mutate(percent = n / sum(n) * 100)
```

```{r}
ggplot(bc, aes(x = Class, fill = Class)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Class balance",
    x = "Diagnosis",
    y = "Count"
  )
```

---

## Exploratory visualization

```{r}
ggplot(bc, aes(x = Cl.thickness, fill = Class)) +
  geom_histogram(alpha = 0.6, bins = 10, position = "identity") +
  theme_minimal() +
  labs(
    title = "Clump thickness by diagnosis",
    x = "Clump thickness",
    y = "Count"
  )
```

```{r}
ggplot(bc, aes(x = Class, y = Cell.size, fill = Class)) +
  geom_boxplot(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Cell size by diagnosis",
    x = "Diagnosis",
    y = "Uniformity of cell size"
  )
```

```{r}
ggplot(bc, aes(x = Cell.size, y = Cell.shape, color = Class)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Cell size vs cell shape",
    x = "Uniformity of cell size",
    y = "Uniformity of cell shape"
  )
```

---

# 7. Why train/test split is necessary

A model can look very good on the same data used to train it.
This can happen because the model has already seen those examples.
If we evaluate only on training data, we answer:
But in real work, we usually care about:
How well will the model work on new unseen samples?

So we split data into:
- **training set**: used to fit model parameters;
- **test set**: kept aside and used only for final evaluation.
This simulates future data.


## Data leakage

Data leakage happens when information from the test set accidentally influences model training.

Examples:

- selecting predictors after looking at test performance;
- scaling using all data before splitting;
- imputing missing values using all data before splitting;
- choosing threshold based on test set repeatedly;
- duplicate samples appearing in both train and test.


## Train/test split

We should preserve class proportions in both sets.  
For simplicity, we do a stratified split manually.

```{r}
set.seed(2026)

bc_train <- bc %>%
  group_by(Class) %>%
  slice_sample(prop = 0.75) %>%
  ungroup()

bc_test <- bc %>%
  anti_join(bc_train, by = names(bc))

bc_train %>% count(Class)
bc_test %>% count(Class)
```

# pandas equivalent:
# train, test = train_test_split(df, test_size=0.25, random_state=42, stratify=df["diagnosis"])

---

# 8. Fit logistic regression

We fit the model using `glm()` with `family = binomial`.

The positive class is the first level of `Class`, which we set to `malignant`.

```{r}
logistic_model <- glm(
  Class ~ Cl.thickness + Cell.size + Cell.shape + Marg.adhesion + Bare.nuclei,
  data = bc_train,
  family = binomial)

summary(logistic_model)
```

# pandas equivalent:
# model = sm.Logit.from_formula(
#     "diagnosis_binary ~ clump_thickness + cell_size + cell_shape + marginal_adhesion + bare_nuclei",
#     data=train
# ).fit()

---

## Coefficients as odds ratios

Logistic regression coefficients are in log-odds units.

To convert them to odds ratios:

```{r}
tidy(logistic_model) %>%
  mutate(odds_ratio = exp(estimate))
```

Interpretation:

- odds ratio above 1 means higher odds of the positive class;
- odds ratio below 1 means lower odds of the positive class;
- the effect is conditional on other variables in the model.

---

# 9. Predicted probabilities

```{r}
test_predictions <- bc_test %>%
  mutate(
    predicted_probability = predict(
      logistic_model,
      newdata = bc_test,
      type = "response"
    )
  )

test_predictions %>%
  select(Class, predicted_probability) %>%
  head()
```

Important:

Since `Class` is ordered as `malignant`, `benign`, the probability is interpreted as probability of `malignant`.

---

## Probability visualization

```{r}
ggplot(test_predictions, aes(x = predicted_probability, y = Class, color = Class)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Predicted probabilities on the test set",
    subtitle = "Dashed line shows threshold = 0.5",
    x = "Predicted probability of malignant diagnosis",
    y = "True class"
  )
```

---

# 10. Thresholds

A logistic regression model returns probabilities.

To convert probabilities into classes, we choose a threshold.

Common default:

$$
P(Y = 1) \geq 0.5 \Rightarrow positive
$$

Here, positive means `malignant`.

```{r}
test_predictions <- test_predictions %>%
  mutate(
    predicted_class = if_else(
      predicted_probability >= 0.5,
      "malignant",
      "benign"
    ),
    predicted_class = factor(predicted_class, levels = c("malignant", "benign"))
  )

test_predictions %>%
  select(Class, predicted_probability, predicted_class) %>%
  head()
```

---

# 11. Confusion matrix

A confusion matrix compares true labels and predicted labels.

For binary classification:

| | Predicted positive | Predicted negative |
|---|---:|---:|
| Actual positive | TP | FN |
| Actual negative | FP | TN |

In this case:

- positive = malignant;
- negative = benign.

```{r}
confusion <- table(
  Truth = test_predictions$Class,
  Prediction = test_predictions$predicted_class
)

confusion
```

---

## Confusion matrix heatmap

```{r}
as.data.frame(confusion) %>%
  ggplot(aes(x = Prediction, y = Truth, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 8) +
  theme_minimal() +
  labs(
    title = "Confusion matrix",
    x = "Predicted class",
    y = "True class",
    fill = "Count"
  )
```

---

# 12. Classification metrics

Let us define the positive class as:

- `malignant`

```{r}
metric_data <- test_predictions %>%
  mutate(
    Class = factor(Class, levels = c("malignant", "benign")),
    predicted_class = factor(predicted_class, levels = c("malignant", "benign"))
  )
```

---

## Accuracy

Accuracy answers:

> What fraction of all predictions are correct?

$$
Accuracy = \frac{TP + TN}{TP + TN + FP + FN}
$$

```{r}
accuracy(metric_data, truth = Class, estimate = predicted_class)
```

Accuracy can be misleading when classes are imbalanced.

---

## Precision

Precision answers:

> When the model predicts malignant, how often is it correct?

$$
Precision = \frac{TP}{TP + FP}
$$

```{r}
precision(metric_data, truth = Class, estimate = predicted_class)
```

---

## Recall / sensitivity

Recall answers:

> Of all truly malignant samples, how many did the model find?

$$
Recall = \frac{TP}{TP + FN}
$$

```{r}
recall(metric_data, truth = Class, estimate = predicted_class)
```

In medical screening, recall is often very important because false negatives can be dangerous.

---

## Specificity

Specificity answers:

> Of all truly benign samples, how many were correctly identified as benign?

$$
Specificity = \frac{TN}{TN + FP}
$$

```{r}
spec(metric_data, truth = Class, estimate = predicted_class)
```

---

## F1-score

F1-score balances precision and recall.

$$
F1 = 2 \cdot \frac{Precision \cdot Recall}{Precision + Recall}
$$

```{r}
f_meas(metric_data, truth = Class, estimate = predicted_class)
```

---

## Metric table

```{r}
classification_metrics <- metric_set(accuracy, precision, recall, spec, f_meas)

classification_metrics(
  metric_data,
  truth = Class,
  estimate = predicted_class
)
```

---

# 13. Threshold tradeoff

Changing the threshold changes the metrics.

Lower threshold:

- more malignant predictions;
- higher recall;
- more false positives.

Higher threshold:

- fewer malignant predictions;
- higher precision;
- more false negatives.

```{r}
threshold_results <- tibble(
  threshold = seq(0.1, 0.9, by = 0.1)
) %>%
  mutate(
    predictions = map(
      threshold,
      ~ test_predictions %>%
        mutate(
          predicted_class = factor(
            if_else(predicted_probability >= .x, "malignant", "benign"),
            levels = c("malignant", "benign")
          ),
          Class = factor(Class, levels = c("malignant", "benign"))
        )
    ),
    accuracy = map_dbl(predictions, ~ accuracy(.x, truth = Class, estimate = predicted_class)$.estimate),
    precision = map_dbl(predictions, ~ precision(.x, truth = Class, estimate = predicted_class)$.estimate),
    recall = map_dbl(predictions, ~ recall(.x, truth = Class, estimate = predicted_class)$.estimate),
    specificity = map_dbl(predictions, ~ spec(.x, truth = Class, estimate = predicted_class)$.estimate),
    f1 = map_dbl(predictions, ~ f_meas(.x, truth = Class, estimate = predicted_class)$.estimate)
  ) %>%
  select(-predictions)

threshold_results
```

```{r}
threshold_results %>%
  pivot_longer(
    cols = c(accuracy, precision, recall, specificity, f1),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = threshold, y = value, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Effect of classification threshold",
    subtitle = "Lower threshold usually increases recall but decreases precision",
    x = "Threshold",
    y = "Metric value",
    color = "Metric"
  )
```

---

# 14. ROC curve and AUC

ROC curve shows performance over all possible thresholds.

It plots:

- true positive rate;
- false positive rate.

AUC summarizes the ROC curve.

AUC interpretation:

- 0.5: random classifier;
- 0.7: acceptable;
- 0.8: good;
- 0.9: excellent;
- 1.0: perfect.

```{r}
roc_obj <- roc(
  response = test_predictions$Class,
  predictor = test_predictions$predicted_probability,
  levels = c("benign", "malignant")
)

auc(roc_obj)
```

```{r}
plot(
  roc_obj,
  col = "steelblue",
  lwd = 3,
  main = "ROC curve for Breast Cancer logistic regression model"
)

abline(a = 0, b = 1, lty = 2, col = "gray")
```

---

## Manual ROC-style threshold table

```{r}
roc_threshold_table <- tibble(
  threshold = seq(0.05, 0.95, by = 0.05)
) %>%
  mutate(
    predicted_class = map(
      threshold,
      ~ if_else(test_predictions$predicted_probability >= .x, "malignant", "benign")
    ),
    tpr = map_dbl(
      predicted_class,
      ~ sum(.x == "malignant" & test_predictions$Class == "malignant") /
        sum(test_predictions$Class == "malignant")
    ),
    fpr = map_dbl(
      predicted_class,
      ~ sum(.x == "malignant" & test_predictions$Class == "benign") /
        sum(test_predictions$Class == "benign")
    )
  ) %>%
  select(-predicted_class)

roc_threshold_table
```



# 15. Comparing training vs test performance

Now we show why evaluating only on training data is risky.

```{r}
train_predictions <- bc_train %>%
  mutate(
    predicted_probability = predict(
      logistic_model,
      newdata = bc_train,
      type = "response"
    ),
    predicted_class = factor(
      if_else(predicted_probability >= 0.5, "malignant", "benign"),
      levels = c("malignant", "benign")
    ),
    Class = factor(Class, levels = c("malignant", "benign"))
  )

train_metrics <- classification_metrics(
  train_predictions,
  truth = Class,
  estimate = predicted_class
) %>%
  mutate(dataset = "train")

test_metrics <- classification_metrics(
  metric_data,
  truth = Class,
  estimate = predicted_class
) %>%
  mutate(dataset = "test")

bind_rows(train_metrics, test_metrics) %>%
  select(dataset, .metric, .estimate) %>%
  arrange(.metric, dataset)
```

Usually, training performance is equal to or better than test performance.

If training performance is much better than test performance, this suggests overfitting.



# Homework: to build the best possible logisitic regression model on wisconsin data


```{r}

```