---
title: "Practical: Gradient Descent and Logistic Regression"
subtitle: "Optimization, classification, thresholds, and model metrics"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
---




# Setup

```{r}
library(tidyverse)
library(broom)
library(yardstick)
library(pROC)
```

# pandas equivalent:
# import pandas as pd
# import numpy as np
# import seaborn as sns
# import matplotlib.pyplot as plt
# import statsmodels.api as sm
# from sklearn.model_selection import train_test_split
# from sklearn.linear_model import LogisticRegression
# from sklearn.metrics import confusion_matrix, accuracy_score, precision_score, recall_score, f1_score, roc_curve, roc_auc_score


# 1. Why do we need gradient descent?

In linear regression, we write:

$$
\hat{y} = \beta_0 + \beta_1x
$$

But how do we find the best values of `beta_0` and `beta_1`?

We need a **loss function**.

For regression, a common loss function is mean squared error:

$$
MSE = \frac{1}{n}\sum(y_i - \hat{y}_i)^2
$$

The model tries to find parameters that make this loss as small as possible.


## Visual idea of a loss function

Let us imagine that we have only one parameter, `theta`.

The loss function is:

$$
J(\theta) = (\theta - 2)^2 + 1
$$

The minimum is at `theta = 2`.

```{r}
loss_data <- tibble(
  theta = seq(-5, 8, length.out = 300),
  loss = (theta - 2)^2 + 1
)

ggplot(loss_data, aes(x = theta, y = loss)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 2, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "A simple loss function",
    subtitle = "The best parameter value is where the loss is minimal",
    x = "Parameter theta",
    y = "Loss"
  )
```

# pandas equivalent:
# theta = np.linspace(-5, 8, 300)
# loss = (theta - 2)**2 + 1
# plt.plot(theta, loss)

---

# 2. Gradient descent intuition

Gradient descent is an iterative optimization method.

The update rule is:

$$
\theta_{new} = \theta_{old} - \eta \nabla J(\theta)
$$

Where:

- `theta` is the parameter;
- `eta` is the learning rate;
- `gradient` is the slope of the loss function;
- each step moves the parameter toward lower loss.

For our loss function:

$$
J(\theta) = (\theta - 2)^2 + 1
$$

the gradient is:

$$
\nabla J(\theta) = 2(\theta - 2)
$$


## Manual gradient descent simulation

```{r}
theta <- -4
learning_rate <- 0.2
n_steps <- 15

gd_path <- tibble(
  step = 0,
  theta = theta,
  loss = (theta - 2)^2 + 1)

for (i in 1:n_steps) {
  gradient <- 2 * (theta - 2)
  theta <- theta - learning_rate * gradient

  gd_path <- bind_rows(
    gd_path,
    tibble(
      step = i,
      theta = theta,
      loss = (theta - 2)^2 + 1
    )
  )
}

gd_path
```

```{r}
ggplot(loss_data, aes(x = theta, y = loss)) +
  geom_line(linewidth = 1.2) +
  geom_point(
    data = gd_path,
    aes(x = theta, y = loss, color = step),
    size = 3
  ) +
  geom_path(
    data = gd_path,
    aes(x = theta, y = loss),
    arrow = arrow(length = unit(0.15, "cm")),
    linewidth = 0.8
  ) +
  theme_minimal() +
  labs(
    title = "Gradient descent path",
    subtitle = "The parameter moves step by step toward the minimum",
    x = "Parameter theta",
    y = "Loss",
    color = "Step"
  )
```

---

## Learning rate effect

The learning rate controls step size.

If it is too small, optimization is slow.  
If it is too large, optimization can overshoot or diverge.

```{r}
simulate_gd <- function(start_theta, learning_rate, n_steps = 15) {
  theta <- start_theta

  out <- tibble(
    step = 0,
    theta = theta,
    loss = (theta - 2)^2 + 1,
    learning_rate = learning_rate
  )

  for (i in 1:n_steps) {
    gradient <- 2 * (theta - 2)
    theta <- theta - learning_rate * gradient

    out <- bind_rows(
      out,
      tibble(
        step = i,
        theta = theta,
        loss = (theta - 2)^2 + 1,
        learning_rate = learning_rate
      )
    )
  }

  out
}

gd_rates <- bind_rows(
  simulate_gd(-4, 0.05),
  simulate_gd(-4, 0.2),
  simulate_gd(-4, 0.9)
)

ggplot(gd_rates, aes(x = step, y = loss, color = as.factor(learning_rate))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Effect of learning rate",
    x = "Gradient descent step",
    y = "Loss",
    color = "Learning rate"
  )
```


# 3. From regression to classification

Linear regression predicts numeric values:

$$
\hat{y} = \beta_0 + \beta_1x
$$

But classification needs class labels.

For example:

- disease: yes/no;
- spam: yes/no;
- survival: yes/no;
- response to treatment: responder/non-responder.

If we use linear regression for binary classes, it can predict invalid values like:

- `-0.2`
- `1.4`

Instead, we want a model that predicts probabilities between 0 and 1.


# 4. Sigmoid function

The sigmoid function converts any real number into a value between 0 and 1.

$$
\sigma(z) = \frac{1}{1 + e^{-z}}
$$

```{r}
sigmoid_data <- tibble(
  z = seq(-10, 10, length.out = 300),
  probability = 1 / (1 + exp(-z))
)

ggplot(sigmoid_data, aes(x = z, y = probability)) +
  geom_line(linewidth = 1.3, color = "steelblue") +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Sigmoid function",
    subtitle = "Any real value is transformed into a probability",
    x = "Linear predictor z",
    y = "Predicted probability"
  )
```

---

# 5. Logistic regression model

Logistic regression has two parts.

First, a linear predictor:

$$
z = \beta_0 + \beta_1x_1 + \beta_2x_2 + ...
$$

Then the sigmoid transformation:

$$
P(Y = 1) = \frac{1}{1 + e^{-z}}
$$

So logistic regression is still a linear model, but it predicts probability.



# 6. Classification dataset

We will use the built-in `mtcars` dataset.

It is small, but useful for teaching.

We will predict whether a car has an automatic or manual transmission.

In `mtcars`:

- `am = 0`: automatic;
- `am = 1`: manual.

```{r}
cars <- mtcars %>%
  rownames_to_column("car_name") %>%
  as_tibble() %>%
  mutate(
    transmission = factor(
      am,
      levels = c(0, 1),
      labels = c("automatic", "manual")
    )
  )

glimpse(cars)
```


## Exploratory visualization

```{r}
ggplot(cars, aes(x = wt, y = mpg, color = transmission)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Cars by transmission type",
    x = "Weight",
    y = "Miles per gallon",
    color = "Transmission"
  )
```

```{r}
ggplot(cars, aes(x = transmission, y = wt, fill = transmission)) +
  geom_boxplot(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Car weight by transmission type",
    x = "Transmission",
    y = "Weight"
  )
```

---

# 7. Train/test split

```{r}
set.seed(42)

train_ids <- sample(seq_len(nrow(cars)), size = 0.75 * nrow(cars))

train_cars <- cars[train_ids, ]
test_cars <- cars[-train_ids, ]

nrow(train_cars)
nrow(test_cars)
```

---

# 8. Fit logistic regression

We fit the model using `glm()` with `family = binomial`.

```{r}
logistic_model <- glm(
  am ~ mpg + wt + hp,
  data = train_cars,
  family = binomial
)

summary(logistic_model)
```

# pandas equivalent:
# model = sm.Logit.from_formula("am ~ mpg + wt + hp", data=train).fit()
# print(model.summary())

---

## Coefficients as odds

Logistic regression coefficients are in log-odds units.

To convert them to odds ratios:

```{r}
tidy(logistic_model) %>%
  mutate(odds_ratio = exp(estimate))
```

Interpretation:

- odds ratio above 1 means higher odds of `am = 1`;
- odds ratio below 1 means lower odds of `am = 1`;
- the effect is conditional on the other variables in the model.

---

# 9. Predicted probabilities

```{r}
test_predictions <- test_cars %>%
  mutate(
    predicted_probability = predict(
      logistic_model,
      newdata = test_cars,
      type = "response"
    )
  )

test_predictions %>%
  select(car_name, transmission, am, predicted_probability)
```

---

## Probability visualization

```{r}
ggplot(test_predictions, aes(x = predicted_probability, y = transmission, color = transmission)) +
  geom_point(size = 4) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Predicted probabilities",
    subtitle = "Dashed line shows threshold = 0.5",
    x = "Predicted probability of manual transmission",
    y = "True class"
  )
```

---

# 10. Thresholds

A logistic regression model returns probabilities.

To convert probabilities into classes, we choose a threshold.

Common default:

$$
P(Y = 1) \geq 0.5 \Rightarrow positive
$$

```{r}
test_predictions <- test_predictions %>%
  mutate(
    predicted_class = if_else(
      predicted_probability >= 0.5,
      1,
      0
    ),
    predicted_transmission = factor(
      predicted_class,
      levels = c(0, 1),
      labels = c("automatic", "manual")
    )
  )

test_predictions %>%
  select(car_name, transmission, predicted_probability, predicted_transmission)
```

---

# 11. Confusion matrix

A confusion matrix compares true labels and predicted labels.

For binary classification:

| | Predicted positive | Predicted negative |
|---|---:|---:|
| Actual positive | TP | FN |
| Actual negative | FP | TN |

```{r}
confusion <- table(
  Truth = test_predictions$transmission,
  Prediction = test_predictions$predicted_transmission
)

confusion
```

---

## Confusion matrix heatmap

```{r}
as.data.frame(confusion) %>%
  ggplot(aes(x = Prediction, y = Truth, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 8) +
  theme_minimal() +
  labs(
    title = "Confusion matrix",
    x = "Predicted class",
    y = "True class",
    fill = "Count"
  )
```

---

# 12. Classification metrics

Let us define the positive class as:

- `manual`

```{r}
metric_data <- test_predictions %>%
  mutate(
    transmission = factor(transmission, levels = c("manual", "automatic")),
    predicted_transmission = factor(predicted_transmission, levels = c("manual", "automatic"))
  )
```

---

## Accuracy

Accuracy answers:

> What fraction of all predictions are correct?

$$
Accuracy = \frac{TP + TN}{TP + TN + FP + FN}
$$

```{r}
accuracy(metric_data, truth = transmission, estimate = predicted_transmission)
```

---

## Precision

Precision answers:

> When the model predicts positive, how often is it correct?

$$
Precision = \frac{TP}{TP + FP}
$$

```{r}
precision(metric_data, truth = transmission, estimate = predicted_transmission)
```

---

## Recall / sensitivity

Recall answers:

> Of all real positives, how many did the model find?

$$
Recall = \frac{TP}{TP + FN}
$$

```{r}
recall(metric_data, truth = transmission, estimate = predicted_transmission)
```

---

## F1-score

F1-score balances precision and recall.

$$
F1 = 2 \cdot \frac{Precision \cdot Recall}{Precision + Recall}
$$

```{r}
f_meas(metric_data, truth = transmission, estimate = predicted_transmission)
```

---

## Metric table

```{r}
classification_metrics <- metric_set(accuracy, precision, recall, f_meas)

classification_metrics(
  metric_data,
  truth = transmission,
  estimate = predicted_transmission
)
```

---

# 13. Threshold tradeoff

Changing the threshold changes the metrics.

Lower threshold:

- more positive predictions;
- higher recall;
- more false positives.

Higher threshold:

- fewer positive predictions;
- higher precision;
- more false negatives.

```{r}
threshold_results <- tibble(
  threshold = seq(0.1, 0.9, by = 0.1)
) %>%
  mutate(
    predictions = map(
      threshold,
      ~ test_predictions %>%
        mutate(
          predicted_transmission = factor(
            if_else(predicted_probability >= .x, "manual", "automatic"),
            levels = c("manual", "automatic")
          ),
          transmission = factor(transmission, levels = c("manual", "automatic"))
        )
    ),
    accuracy = map_dbl(predictions, ~ accuracy(.x, truth = transmission, estimate = predicted_transmission)$.estimate),
    precision = map_dbl(predictions, ~ precision(.x, truth = transmission, estimate = predicted_transmission)$.estimate),
    recall = map_dbl(predictions, ~ recall(.x, truth = transmission, estimate = predicted_transmission)$.estimate),
    f1 = map_dbl(predictions, ~ f_meas(.x, truth = transmission, estimate = predicted_transmission)$.estimate)
  ) %>%
  select(-predictions)

threshold_results
```

```{r}
threshold_results %>%
  pivot_longer(
    cols = c(accuracy, precision, recall, f1),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = threshold, y = value, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Effect of classification threshold",
    x = "Threshold",
    y = "Metric value",
    color = "Metric"
  )
```

---

# 14. ROC curve and AUC

ROC curve shows performance over all possible thresholds.

It plots:

- true positive rate;
- false positive rate.

AUC summarizes the ROC curve.

AUC interpretation:

- 0.5: random classifier;
- 0.7: acceptable;
- 0.8: good;
- 0.9: excellent;
- 1.0: perfect.

```{r}
roc_obj <- roc(
  response = test_predictions$am,
  predictor = test_predictions$predicted_probability
)

auc(roc_obj)
```

```{r}
plot(
  roc_obj,
  col = "steelblue",
  lwd = 3,
  main = "ROC curve for logistic regression model"
)

abline(a = 0, b = 1, lty = 2, col = "gray")
```

---

## Manual ROC-style threshold table

```{r}
roc_threshold_table <- tibble(
  threshold = seq(0.05, 0.95, by = 0.05)
) %>%
  mutate(
    predicted_class = map(
      threshold,
      ~ if_else(test_predictions$predicted_probability >= .x, 1, 0)
    ),
    tpr = map_dbl(
      predicted_class,
      ~ sum(.x == 1 & test_predictions$am == 1) / sum(test_predictions$am == 1)
    ),
    fpr = map_dbl(
      predicted_class,
      ~ sum(.x == 1 & test_predictions$am == 0) / sum(test_predictions$am == 0)
    )
  ) %>%
  select(-predicted_class)

roc_threshold_table
```

```{r}
ggplot(roc_threshold_table, aes(x = fpr, y = tpr)) +
  geom_line(linewidth = 1.2) +
  geom_point(aes(color = threshold), size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Manual ROC curve illustration",
    subtitle = "Each point corresponds to a different threshold",
    x = "False positive rate",
    y = "True positive rate",
    color = "Threshold"
  )
```

---

# 15. Class imbalance

Accuracy can be very misleading when one class is much more common.

Example:

```{r}
imbalanced_example <- tibble(
  truth = c(rep("negative", 95), rep("positive", 5)),
  prediction = rep("negative", 100)
)

imbalanced_example %>%
  count(truth, prediction)
```

The model predicts every case as negative.

```{r}
imbalanced_example <- imbalanced_example %>%
  mutate(
    truth = factor(truth, levels = c("positive", "negative")),
    prediction = factor(prediction, levels = c("positive", "negative"))
  )

accuracy(imbalanced_example, truth = truth, estimate = prediction)
recall(imbalanced_example, truth = truth, estimate = prediction)
```

Interpretation:

- accuracy is high;
- recall for positive class is zero;
- the model is useless for finding positives.

---

# 16. In-class challenge

Using the `cars` dataset:

1. Fit a logistic regression model predicting `am`.
2. Use at least three predictors.
3. Predict probabilities on test data.
4. Try thresholds `0.3`, `0.5`, and `0.7`.
5. Calculate:
   - accuracy;
   - precision;
   - recall;
   - F1-score;
   - AUC.
6. Decide which threshold you would use.

```{r}
# Your code here
```

Discussion:

- If this were a medical screening task, would you prefer high recall or high precision?
- If this were an expensive confirmatory diagnostic test, would your preferred threshold change?

---

# 17. Homework: Logistic regression on synthetic medical data

For homework, students will use a synthetic medical screening dataset.

The target variable is:

- `diabetes`

The predictors are:

- `age`;
- `bmi`;
- `glucose`;
- `blood_pressure`;
- `exercise_hours`;
- `smoking`;
- `family_history`;
- `noise_feature`.

The dataset contains:

- class imbalance;
- missing values;
- correlated predictors;
- one irrelevant feature;
- some noisy observations.

---

## Generate homework dataset

```{r}
set.seed(2030)

n <- 250

homework_medical <- tibble(
  patient_id = 1:n,
  age = round(rnorm(n, mean = 50, sd = 13)),
  bmi = round(rnorm(n, mean = 27, sd = 5), 1),
  glucose = round(rnorm(n, mean = 105, sd = 25), 1),
  blood_pressure = round(rnorm(n, mean = 125, sd = 15)),
  exercise_hours = round(pmax(rnorm(n, mean = 3, sd = 2), 0), 1),
  smoking = sample(c("yes", "no"), n, replace = TRUE, prob = c(0.28, 0.72)),
  family_history = sample(c("yes", "no"), n, replace = TRUE, prob = c(0.35, 0.65)),
  noise_feature = rnorm(n)
) %>%
  mutate(
    glucose = glucose + 0.35 * (bmi - mean(bmi)),
    linear_risk =
      -7.5 +
      0.045 * age +
      0.09 * bmi +
      0.035 * glucose +
      0.012 * blood_pressure -
      0.18 * exercise_hours +
      if_else(smoking == "yes", 0.55, 0) +
      if_else(family_history == "yes", 0.75, 0),
    diabetes_probability = 1 / (1 + exp(-linear_risk)),
    diabetes = rbinom(n, size = 1, prob = diabetes_probability),
    diabetes = factor(diabetes, levels = c(0, 1), labels = c("no", "yes"))
  ) %>%
  select(-linear_risk, -diabetes_probability) %>%
  mutate(
    glucose = if_else(patient_id %in% c(8, 44, 101, 199), NA_real_, glucose),
    bmi = if_else(patient_id %in% c(12, 88, 150), NA_real_, bmi),
    smoking = if_else(patient_id %in% c(33, 120), NA_character_, smoking),
    glucose = if_else(patient_id %in% c(20, 77), glucose + 80, glucose)
  )

homework_medical
```

---

## Homework task 1: Inspect data

Check:

- structure;
- missing values;
- target class balance;
- summary statistics;
- distributions.

```{r}
# Your code here
```

---

## Homework task 2: Clean data

Choose how to handle missing values.

Possible approaches:

- remove rows with missing values;
- median imputation for numeric columns;
- mode imputation for categorical columns.

```{r}
# Your code here
```

---

## Homework task 3: Exploratory plots

Create:

- histogram of glucose;
- boxplot of BMI by diabetes status;
- scatter plot of glucose vs BMI colored by diabetes;
- bar chart of diabetes by family history.

```{r}
# Your code here
```

---

## Homework task 4: Train/test split

Split data into training and test subsets.

```{r}
# Your code here
```

---

## Homework task 5: Fit candidate logistic models

Build at least three models.

Examples:

```{r}
# model_1 <- glm(diabetes ~ glucose, data = train_data, family = binomial)
# model_2 <- glm(diabetes ~ glucose + bmi + age, data = train_data, family = binomial)
# model_3 <- glm(diabetes ~ glucose + bmi + age + exercise_hours + smoking + family_history,
#                data = train_data, family = binomial)
```

---

## Homework task 6: Evaluate model

For your best model:

1. predict probabilities;
2. apply threshold 0.5;
3. compute confusion matrix;
4. calculate accuracy, precision, recall, F1-score;
5. calculate ROC AUC;
6. plot ROC curve.

```{r}
# Your code here
```

---

## Homework task 7: Threshold decision

Compare thresholds:

- `0.3`;
- `0.5`;
- `0.7`.

Questions:

1. Which threshold gives the highest recall?
2. Which threshold gives the highest precision?
3. If this is a screening tool, which threshold would you choose?
4. If this is a confirmatory diagnostic tool, which threshold would you choose?

```{r}
# Your code here
```

---

## Homework final answer

Write a short conclusion:

1. Which model did you choose?
2. Which predictors were most useful?
3. What threshold did you choose?
4. Why?
5. What are the limitations of the model?
6. What additional data would improve the model?

---

# Suggested submission format

Students should submit:

1. data cleaning code;
2. exploratory visualizations;
3. at least three logistic regression models;
4. model evaluation table;
5. confusion matrix;
6. ROC curve;
7. final threshold decision;
8. short written interpretation.
