---
title: "Practical: Tidy Data and Data Manipulation with tidyverse"
subtitle: "50-minute beginner-friendly practical"
author: ""
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## Learning goals

By the end of this practical, students should be able to:

1. Explain what **tidy data** means.
2. Use core `dplyr` verbs: `select()`, `filter()`, `mutate()`, `summarise()`, `group_by()`, `arrange()`.
3. Reshape data with `pivot_longer()` and `pivot_wider()`.
4. Combine tables using joins.
5. Translate the main operations conceptually to **pandas**.

Recommended dataset: **Palmer Penguins**. It is small, clean enough for beginners, but realistic enough to demonstrate missing values, categories, grouping, and reshaping.

Approximate timing:

| Section | Time |
|---|---:|
| Setup and tidy data concept | 5 min |
| Inspecting and selecting data | 7 min |
| Filtering and arranging | 7 min |
| Mutating new columns | 8 min |
| Grouping and summarising | 8 min |
| Pivoting | 8 min |
| Joins | 5 min |
| Mini exercise | 2 min |

## 1. Setup

```{r}
# Install once if needed:
# install.packages("tidyverse")
# install.packages("palmerpenguins")

library(tidyverse)
library(palmerpenguins)

penguins <- penguins
```

```{r}
# Quick look at the data
penguins
```

```{r}
glimpse(penguins)
```

```{r}
summary(penguins)
```

```{r}
# pandas equivalent:
# import pandas as pd
# df.head()
# df.info()
# df.describe(include="all")
```

## 2. What is tidy data?

A dataset is usually called **tidy** when:

1. Each variable is a column.
2. Each observation is a row.
3. Each value is a cell.

In `penguins`, each row is one penguin. Columns such as `species`, `island`, `bill_length_mm`, and `body_mass_g` are variables.

```{r}
# Check column names
names(penguins)
```

```{r}
# Number of rows and columns
nrow(penguins)
ncol(penguins)
```

```{r}
# pandas equivalent:
# df.columns
# df.shape
```

## 3. Selecting columns with `select()`

Use `select()` when you want to keep only specific columns.

```{r}
penguins_basic <- penguins %>%
  select(species, island, sex, body_mass_g)

penguins_basic
```

```{r}
# Select columns by pattern
penguins_bill <- penguins %>%
  select(species, island, starts_with("bill"))

penguins_bill
```

```{r}
# Remove a column
penguins_no_year <- penguins %>%
  select(-year)

penguins_no_year
```

```{r}
# pandas equivalent:
# df[["species", "island", "sex", "body_mass_g"]]
# df.filter(regex="^bill")
# df.drop(columns=["year"])
```

### Exercise 1

Create a table containing only `species`, `island`, `flipper_length_mm`, and `body_mass_g`.

```{r}
# Your code here
```

## 4. Filtering rows with `filter()`

Use `filter()` to keep rows that satisfy a condition.

```{r}
# Keep only Adelie penguins
adelie <- penguins %>%
  filter(species == "Adelie")

adelie
```

```{r}
# Keep penguins with body mass above 5000 g
large_penguins <- penguins %>%
  filter(body_mass_g > 5000)

large_penguins
```

```{r}
# Multiple conditions: Gentoo penguins from Biscoe island
biscoe_gentoo <- penguins %>%
  filter(species == "Gentoo", island == "Biscoe")

biscoe_gentoo
```

```{r}
# OR condition
adelie_or_chinstrap <- penguins %>%
  filter(species == "Adelie" | species == "Chinstrap")

adelie_or_chinstrap
```

```{r}
# More compact OR condition
adelie_or_chinstrap <- penguins %>%
  filter(species %in% c("Adelie", "Chinstrap"))

adelie_or_chinstrap
```

```{r}
# Remove missing body mass values
penguins_mass_known <- penguins %>%
  filter(!is.na(body_mass_g))

penguins_mass_known
```

```{r}
# pandas equivalent:
# df[df["species"] == "Adelie"]
# df[df["body_mass_g"] > 5000]
# df[(df["species"] == "Gentoo") & (df["island"] == "Biscoe")]
# df[df["species"].isin(["Adelie", "Chinstrap"])]
# df[df["body_mass_g"].notna()]
```

### Exercise 2

Keep only penguins from `Dream` island with known sex.

```{r}
# Your code here
```

## 5. Sorting rows with `arrange()`

Use `arrange()` to order rows.

```{r}
# Lightest penguins first
penguins %>%
  arrange(body_mass_g)
```

```{r}
# Heaviest penguins first
penguins %>%
  arrange(desc(body_mass_g))
```

```{r}
# Sort by species, then by body mass within species
penguins %>%
  arrange(species, desc(body_mass_g))
```

```{r}
# pandas equivalent:
# df.sort_values("body_mass_g")
# df.sort_values("body_mass_g", ascending=False)
# df.sort_values(["species", "body_mass_g"], ascending=[True, False])
```

## 6. Creating new columns with `mutate()`

Use `mutate()` to add or modify columns.

```{r}
# Convert body mass from grams to kilograms
penguins2 <- penguins %>%
  mutate(body_mass_kg = body_mass_g / 1000)

penguins2 %>%
  select(species, body_mass_g, body_mass_kg)
```

```{r}
# Create a ratio between bill length and bill depth
penguins2 <- penguins2 %>%
  mutate(bill_ratio = bill_length_mm / bill_depth_mm)

penguins2 %>%
  select(species, bill_length_mm, bill_depth_mm, bill_ratio)
```

```{r}
# Create categorical variable using case_when()
penguins2 <- penguins2 %>%
  mutate(
    mass_category = case_when(
      body_mass_g < 3500 ~ "small",
      body_mass_g >= 3500 & body_mass_g < 5000 ~ "medium",
      body_mass_g >= 5000 ~ "large",
      TRUE ~ NA_character_
    )
  )

penguins2 %>%
  select(species, body_mass_g, mass_category)
```

```{r}
# pandas equivalent:
# df["body_mass_kg"] = df["body_mass_g"] / 1000
# df["bill_ratio"] = df["bill_length_mm"] / df["bill_depth_mm"]
# import numpy as np
# df["mass_category"] = np.select(
#     [df["body_mass_g"] < 3500,
#      (df["body_mass_g"] >= 3500) & (df["body_mass_g"] < 5000),
#      df["body_mass_g"] >= 5000],
#     ["small", "medium", "large"],
#     default=np.nan
# )
```

### Exercise 3

Create a new column called `flipper_length_cm` by converting `flipper_length_mm` to centimeters.

```{r}
# Your code here
```

## 7. Grouping and summarising

Use `group_by()` together with `summarise()` to calculate statistics by group.

```{r}
# Mean body mass by species
penguins %>%
  group_by(species) %>%
  summarise(mean_body_mass = mean(body_mass_g, na.rm = TRUE))
```

```{r}
# Multiple summaries by species
species_summary <- penguins %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    sd_body_mass = sd(body_mass_g, na.rm = TRUE),
    mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE)
  )

species_summary
```

```{r}
# Group by two variables
penguins %>%
  group_by(species, sex) %>%
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# Count is a shortcut for group_by() + summarise(n = n())
penguins %>%
  count(species, island)
```

```{r}
# pandas equivalent:
# df.groupby("species")["body_mass_g"].mean()
# df.groupby("species").agg(
#     n=("species", "size"),
#     mean_body_mass=("body_mass_g", "mean"),
#     sd_body_mass=("body_mass_g", "std"),
#     mean_flipper_length=("flipper_length_mm", "mean")
# ).reset_index()
# df.groupby(["species", "sex"]).agg(n=("species", "size"), mean_body_mass=("body_mass_g", "mean")).reset_index()
# df.groupby(["species", "island"]).size().reset_index(name="n")
```

### Exercise 4

Calculate the mean bill length for each species and island combination.

```{r}
# Your code here
```

## 8. Using pipes for readable workflows

The pipe `%>%` reads as "then".

Example: take the penguins dataset, then remove missing body mass values, then group by species, then calculate mean body mass, then sort the result.

```{r}
penguins %>%
  filter(!is.na(body_mass_g)) %>%
  group_by(species) %>%
  summarise(mean_body_mass = mean(body_mass_g), .groups = "drop") %>%
  arrange(desc(mean_body_mass))
```

```{r}
# pandas equivalent:
# (df.dropna(subset=["body_mass_g"])
#    .groupby("species", as_index=False)
#    .agg(mean_body_mass=("body_mass_g", "mean"))
#    .sort_values("mean_body_mass", ascending=False))
```

## 9. Pivoting: long and wide data

Sometimes data need to be reshaped.

### 9.1 From wide to long with `pivot_longer()`

Let us keep several measurement columns and convert them into a longer format.

```{r}
penguins_measurements <- penguins %>%
  select(species, island, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

penguins_measurements
```

```{r}
penguins_long <- penguins_measurements %>%
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement",
    values_to = "value"
  )

penguins_long
```

```{r}
# pandas equivalent:
# long = df.melt(
#     id_vars=["species", "island"],
#     value_vars=["bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"],
#     var_name="measurement",
#     value_name="value"
# )
```

Now we can summarise all measurements in the same workflow.

```{r}
penguins_long %>%
  group_by(species, measurement) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")
```

### 9.2 From long to wide with `pivot_wider()`

First, create a summary table.

```{r}
summary_long <- penguins_long %>%
  group_by(species, measurement) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")

summary_long
```

Now convert it to a wider table.

```{r}
summary_wide <- summary_long %>%
  pivot_wider(
    names_from = measurement,
    values_from = mean_value
  )

summary_wide
```

```{r}
# pandas equivalent:
# summary_wide = summary_long.pivot(
#     index="species",
#     columns="measurement",
#     values="mean_value"
# ).reset_index()
```

### Exercise 5

Create a long table containing `species`, `sex`, `bill_length_mm`, and `bill_depth_mm`. The final columns should be `species`, `sex`, `bill_measurement`, and `value`.

```{r}
# Your code here
```

## 10. Joins

Joins combine two tables using one or more key columns.

We will create a small lookup table with ecological notes for each species.

```{r}
species_info <- tibble(
  species = c("Adelie", "Chinstrap", "Gentoo"),
  genus = c("Pygoscelis", "Pygoscelis", "Pygoscelis"),
  typical_region = c("Antarctic coast", "Subantarctic islands", "Subantarctic islands")
)

species_info
```

### 10.1 `left_join()`

`left_join()` keeps all rows from the left table and adds matching information from the right table.

```{r}
penguins_with_info <- penguins %>%
  left_join(species_info, by = "species")

penguins_with_info %>%
  select(species, island, genus, typical_region)
```

```{r}
# pandas equivalent:
# df_with_info = df.merge(species_info, on="species", how="left")
```

### 10.2 Other join types

```{r}
# inner_join(): keep only matching rows from both tables
penguins %>%
  inner_join(species_info, by = "species")
```

```{r}
# anti_join(): rows in penguins with no match in species_info
penguins %>%
  anti_join(species_info, by = "species")
```

```{r}
# pandas equivalent:
# inner = df.merge(species_info, on="species", how="inner")
# anti = df[~df["species"].isin(species_info["species"])]
```

### Exercise 6

Create an island information table with columns `island` and `island_type`, then join it to `penguins`.

```{r}
# Example starting point:
island_info <- tibble(
  island = c("Biscoe", "Dream", "Torgersen"),
  island_type = c("large island", "small island", "small island")
)

# Your join code here
```

## 11. Mini-project: one full tidyverse pipeline

Task: Create a table showing, for each species and sex:

1. Number of penguins.
2. Mean body mass in kilograms.
3. Mean flipper length in centimeters.
4. Only groups with at least 10 observations.
5. Sort from highest to lowest mean body mass.

```{r}
mini_project_result <- penguins %>%
  filter(!is.na(sex), !is.na(body_mass_g), !is.na(flipper_length_mm)) %>%
  mutate(
    body_mass_kg = body_mass_g / 1000,
    flipper_length_cm = flipper_length_mm / 10
  ) %>%
  group_by(species, sex) %>%
  summarise(
    n = n(),
    mean_body_mass_kg = mean(body_mass_kg),
    mean_flipper_length_cm = mean(flipper_length_cm),
    .groups = "drop"
  ) %>%
  filter(n >= 10) %>%
  arrange(desc(mean_body_mass_kg))

mini_project_result
```

```{r}
# pandas equivalent:
# result = (
#     df.dropna(subset=["sex", "body_mass_g", "flipper_length_mm"])
#       .assign(
#           body_mass_kg=lambda x: x["body_mass_g"] / 1000,
#           flipper_length_cm=lambda x: x["flipper_length_mm"] / 10
#       )
#       .groupby(["species", "sex"], as_index=False)
#       .agg(
#           n=("species", "size"),
#           mean_body_mass_kg=("body_mass_kg", "mean"),
#           mean_flipper_length_cm=("flipper_length_cm", "mean")
#       )
#       .query("n >= 10")
#       .sort_values("mean_body_mass_kg", ascending=False)
# )
```

## 12. Optional quick plot

Although this practical focuses on data manipulation, a quick plot helps show why tidy data is useful.

```{r}
penguins %>%
  filter(!is.na(body_mass_g), !is.na(species)) %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Body mass distribution by penguin species",
    x = "Species",
    y = "Body mass, g"
  ) +
  theme(legend.position = "none")
```

```{r}
# pandas/seaborn equivalent:
# import seaborn as sns
# import matplotlib.pyplot as plt
# sns.boxplot(data=df, x="species", y="body_mass_g", hue="species", legend=False)
# plt.title("Body mass distribution by penguin species")
# plt.show()
```

## 13. Summary of key tidyverse verbs

| Verb | Purpose | pandas equivalent |
|---|---|---|
| `select()` | Keep/remove columns | `df[[...]]`, `df.drop(columns=...)` |
| `filter()` | Keep rows by condition | `df[condition]`, `.query()` |
| `mutate()` | Create/modify columns | `df.assign()`, `df["new"] = ...` |
| `arrange()` | Sort rows | `df.sort_values()` |
| `group_by()` + `summarise()` | Grouped statistics | `df.groupby().agg()` |
| `count()` | Count rows by group | `df.groupby(...).size()` |
| `pivot_longer()` | Wide to long | `df.melt()` |
| `pivot_wider()` | Long to wide | `df.pivot()` / `df.pivot_table()` |
| `left_join()` | Add matching columns | `df.merge(..., how="left")` |

## 14. Homework / extension

Using the same dataset:

1. Find the heaviest penguin for each species.
2. Calculate mean body mass by year and species.
3. Create a long-format table with all numeric measurements.
4. Join a manually created table containing species-level notes.
5. Make one plot from a summarised table.

```{r}
# Space for homework code
```
