## 1. Setup

```{r}
# install.packages("tidyverse")
# install.packages("palmerpenguins")

library(tidyverse)
library(palmerpenguins)

penguins <- penguins
```

```{r}
penguins
```

```{r}
glimpse(penguins)
```

```{r}
summary(penguins)
```

```{r}
# pandas equivalent:
# import pandas as pd
# df.head()
# df.info()
# df.describe(include="all")
```

## 2. What is tidy data?

A dataset is usually called **tidy** when:

1. Each variable is a column.
2. Each observation is a row.
3. Each value is a cell.

In `penguins`, each row is one penguin. Columns such as `species`, `island`, `bill_length_mm`, and `body_mass_g` are variables.

```{r}
# Check column names
names(penguins)
```

```{r}
# Number of rows and columns
nrow(penguins)
ncol(penguins)
```

```{r}
# pandas equivalent:
# df.columns
# df.shape
```

## 3. Selecting columns with `select()`

Use `select()` when you want to keep only specific columns.

```{r}
penguins_basic <- penguins %>%select(species, island, sex, body_mass_g)

penguins_basic
```

```{r}
# Select columns by pattern
penguins_bill <- penguins %>%
  select(species, island, starts_with("bill"))

penguins_bill
```

```{r}
# Remove a column
penguins_no_year <- penguins %>%select(-year)

penguins_no_year
```

```{r}
# pandas equivalent:
# df[["species", "island", "sex", "body_mass_g"]]
# df.filter(regex="^bill")
# df.drop(columns=["year"])
```

### Exercise 1

Create a table containing only `species`, `island`, `flipper_length_mm`, and `body_mass_g`.

```{r}

```

## 4. Filtering rows with `filter()`

Use `filter()` to keep rows that satisfy a condition.

```{r}
# Keep only Adelie penguins
adelie <- penguins %>%
  filter(species == "Adelie")

adelie
```

```{r}
# Keep penguins with body mass above 5000 g
large_penguins <- penguins %>%
  filter(body_mass_g > 5000)

large_penguins
```

```{r}
# Multiple conditions: Gentoo penguins from Biscoe island
biscoe_gentoo <- penguins %>%
  filter(species == "Gentoo", island == "Biscoe")

biscoe_gentoo
```

```{r}
# OR condition
adelie_or_chinstrap <- penguins %>%
  filter(species == "Adelie" | species == "Chinstrap")

adelie_or_chinstrap
```

```{r}
# More compact OR condition
adelie_or_chinstrap <- penguins %>%
  filter(species %in% c("Adelie", "Chinstrap"))

adelie_or_chinstrap
```

```{r}
# Remove missing body mass values
penguins_mass_known <- penguins %>% drop_na()

penguins_mass_known
```

```{r}
# pandas equivalent:
# df[df["species"] == "Adelie"]
# df[df["body_mass_g"] > 5000]
# df[(df["species"] == "Gentoo") & (df["island"] == "Biscoe")]
# df[df["species"].isin(["Adelie", "Chinstrap"])]
# df[df["body_mass_g"].notna()]
```


Keep only penguins from `Dream` island.

```{r}
# Your code here
```

## 5. Sorting rows with `arrange()`

Use `arrange()` to order rows.

```{r}
# Lightest penguins first
penguins %>%
  arrange(body_mass_g)
```

```{r}
# Heaviest penguins first
penguins %>%
  arrange(desc(body_mass_g))
```

```{r}
# Sort by species, then by body mass within species
penguins %>%
  arrange(species, desc(body_mass_g))
```

```{r}
# pandas equivalent:
# df.sort_values("body_mass_g")
# df.sort_values("body_mass_g", ascending=False)
# df.sort_values(["species", "body_mass_g"], ascending=[True, False])
```

## 6. Creating new columns with `mutate()`

Use `mutate()` to add or modify columns.

```{r}
# Convert body mass from grams to kilograms
penguins2 <- penguins %>%
  mutate(body_mass_kg = body_mass_g / 1000)

penguins2 %>%
  select(species, body_mass_g, body_mass_kg)
```

```{r}
# Create a ratio between bill length and bill depth
penguins2 <- penguins2 %>%
  mutate(bill_ratio = bill_length_mm / bill_depth_mm)

penguins2 %>%
  select(species, bill_length_mm, bill_depth_mm, bill_ratio)
```

```{r}
# Create categorical variable using case_when()
penguins2 <- penguins2 %>%
  mutate(
    mass_category = case_when(
      body_mass_g < 3500 ~ "small",
      body_mass_g >= 3500 & body_mass_g < 5000 ~ "medium",
      body_mass_g >= 5000 ~ "large",
      TRUE ~ NA_character_))

penguins2 %>%
  select(species, body_mass_g, mass_category)
```

```{r}
# pandas equivalent:
# df["body_mass_kg"] = df["body_mass_g"] / 1000
# df["bill_ratio"] = df["bill_length_mm"] / df["bill_depth_mm"]
# import numpy as np
# df["mass_category"] = np.select(
#     [df["body_mass_g"] < 3500,
#      (df["body_mass_g"] >= 3500) & (df["body_mass_g"] < 5000),
#      df["body_mass_g"] >= 5000],
#     ["small", "medium", "large"],
#     default=np.nan
# )
```

### Exercise 3

Create a new column called `flipper_length_cm` by converting `flipper_length_mm` to centimeters.

```{r}
# Your code here
```

## 7. Grouping and summarising

Use `group_by()` together with `summarise()` to calculate statistics by group.

```{r}
# Mean body mass by species
penguins %>%
  group_by(species) %>%
  summarise(mean_body_mass = mean(body_mass_g, na.rm = TRUE))
```

```{r}
# Multiple summaries by species
species_summary <- penguins %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    sd_body_mass = sd(body_mass_g, na.rm = TRUE),
    mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE))

species_summary
```

```{r}
# Group by two variables
penguins %>%
  group_by(species, sex) %>%
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE))
```

```{r}
# Count is a shortcut for group_by() + summarise(n = n())
penguins %>%
  count(species, island)
```

```{r}
# pandas equivalent:
# df.groupby("species")["body_mass_g"].mean()
# df.groupby("species").agg(
#     n=("species", "size"),
#     mean_body_mass=("body_mass_g", "mean"),
#     sd_body_mass=("body_mass_g", "std"),
#     mean_flipper_length=("flipper_length_mm", "mean")
# ).reset_index()
# df.groupby(["species", "sex"]).agg(n=("species", "size"), mean_body_mass=("body_mass_g", "mean")).reset_index()
# df.groupby(["species", "island"]).size().reset_index(name="n")
```

### Exercise 4

Calculate the mean bill length for each species and island combination.

```{r}
# Your code here
```


## 9. Pivoting: long and wide data

Sometimes data need to be reshaped.

### 9.1 From wide to long with `pivot_longer()`

Let us keep several measurement columns and convert them into a longer format.

```{r}
penguins_measurements <- penguins %>%
  select(species, island, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

penguins_measurements
```

```{r}
penguins_long <- penguins_measurements %>%
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement",
    values_to = "value")

penguins_long
```

```{r}
# pandas equivalent:
# long = df.melt(
#     id_vars=["species", "island"],
#     value_vars=["bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"],
#     var_name="measurement",
#     value_name="value"
# )
```

Now we can summarise all measurements in the same workflow.

```{r}
penguins_long %>%
  group_by(species, measurement) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")
```

### 9.2 From long to wide with `pivot_wider()`

First, create a summary table.

```{r}
summary_long <- penguins_long %>%
  group_by(species, measurement) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")

summary_long
```

Now convert it to a wider table.

```{r}
summary_wide <- summary_long %>%
  pivot_wider(
    names_from = measurement,
    values_from = mean_value
  )

summary_wide
```

```{r}
# pandas equivalent:
# summary_wide = summary_long.pivot(
#     index="species",
#     columns="measurement",
#     values="mean_value"
# ).reset_index()
```


## 10. Joins

Joins combine two tables using one or more key columns.

We will create a small lookup table with ecological notes for each species.

```{r}
species_info <- tibble(
  species = c("Adelie", "Chinstrap", "Gentoo"),
  genus = c("Pygoscelis", "Pygoscelis", "Pygoscelis"),
  typical_region = c("Antarctic coast", "Subantarctic islands", "Subantarctic islands"))

species_info
```

### 10.1 `left_join()`

`left_join()` keeps all rows from the left table and adds matching information from the right table.

```{r}
penguins_with_info <- penguins %>%
  left_join(species_info, by = "species")

penguins_with_info %>%
  select(species, island, genus, typical_region)
```

```{r}
# pandas equivalent:
# df_with_info = df.merge(species_info, on="species", how="left")
```

### 10.2 Other join types

```{r}
# inner_join(): keep only matching rows from both tables
penguins %>%
  inner_join(species_info, by = "species")
```

```{r}
# anti_join(): rows in penguins with no match in species_info
penguins %>%
  anti_join(species_info, by = "species")
```

```{r}
# pandas equivalent:
# inner = df.merge(species_info, on="species", how="inner")
# anti = df[~df["species"].isin(species_info["species"])]
```


```{r}
penguins %>%
  filter(!is.na(body_mass_g), !is.na(species)) %>%
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Body mass distribution by penguin species",
    x = "Species",
    y = "Body mass, g"
  ) +
  theme(legend.position = "none")
```

```{r}
# pandas/seaborn equivalent:
# import seaborn as sns
# import matplotlib.pyplot as plt
# sns.boxplot(data=df, x="species", y="body_mass_g", hue="species", legend=False)
# plt.title("Body mass distribution by penguin species")
# plt.show()
```

## 13. Summary of key tidyverse verbs

| Verb | Purpose | pandas equivalent |
|---|---|---|
| `select()` | Keep/remove columns | `df[[...]]`, `df.drop(columns=...)` |
| `filter()` | Keep rows by condition | `df[condition]`, `.query()` |
| `mutate()` | Create/modify columns | `df.assign()`, `df["new"] = ...` |
| `arrange()` | Sort rows | `df.sort_values()` |
| `group_by()` + `summarise()` | Grouped statistics | `df.groupby().agg()` |
| `count()` | Count rows by group | `df.groupby(...).size()` |
| `pivot_longer()` | Wide to long | `df.melt()` |
| `pivot_wider()` | Long to wide | `df.pivot()` / `df.pivot_table()` |
| `left_join()` | Add matching columns | `df.merge(..., how="left")` |




# Homework: Cleaning and Tidying Synthetic Messy Penguin Data

In this homework, you will work with a deliberately messy version of the `penguins` dataset.

The goal is to practice the full tidyverse workflow:

1. Generate or load messy data
2. Inspect problems
3. Clean missing values and inconsistent labels
4. Filter rows
5. Mutate new variables
6. Pivot from wide to long format
7. Join with another table
8. Summarise grouped data
9. Arrange/order the final result

---






## Homework Task

Clean and analyse the messy dataset.

## Read daata

```{r}
data_penquins<-read_csv()
metada_penquins<-read_csv()
```


### Step 1: Inspect the data

Check:

- number of rows and columns
- column names
- missing values
- unique values in `species`, `island`, `sex`, and `observation_quality`

```{r}

```


### Step 2: Clean categorical variables

Create a new object called `penguins_cleaned_homework`.

Recommended operations:

- remove extra spaces using `str_trim()`
- standardise text case using `str_to_title()` or `str_to_lower()`
- make sex values consistent
- remove rows with missing values in important columns

```{r}

```


### Step 3: Filter useful observations

Keep only:

- high or medium quality observations
- penguins with body mass above 3000 g
- penguins from Biscoe or Dream island

```{r}

```



### Step 4: Add new variables with `mutate()`

Add:

- `body_mass_kg`
- `bill_ratio = bill_length_mm / bill_depth_mm`
- `large_penguin`, where body mass above 4500 g is `"large"` and otherwise `"regular"`

```{r}

```



### Step 5: Pivot to long format

Convert measurement columns into a tidy long table.

Use these columns:

- `bill_length_mm`
- `bill_depth_mm`
- `flipper_length_mm`
- `body_mass_g`


```{r}

```

### Step 6: Join with island metadata

Join the cleaned penguin data with `island_metadata`.

```{r}

```


### Step 7: Group, summarise, and order

Calculate summary statistics by:

- species
- island
- sex
- protected_area

Calculate:

- number of observations
- mean body mass
- median flipper length
- mean bill ratio

```{r}

```


## Expected Output

Result should contain 

1. A cleaned dataset
2. A filtered dataset
3. A long-format dataset created with `pivot_longer()`
4. A joined dataset
5. A grouped summary table ordered by mean body mass



