---
title: "Practical: KNN, SVM, and K-means Clustering"
subtitle: "Distance-based models, margins, kernels, and unsupervised clustering"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
---


```{r}
library(tidyverse)
library(yardstick)
library(pROC)
library(class)
library(e1071)
```

If packages are missing, install once:

```r
install.packages(c("class", "e1071", "yardstick", "pROC"))
```

# pandas equivalent:
# from sklearn.neighbors import KNeighborsClassifier
# from sklearn.svm import SVC
# from sklearn.cluster import KMeans
# from sklearn.preprocessing import StandardScaler

---

# 1. Where are we in the course?

So far we discussed:

- tidy data and data manipulation;
- linear regression;
- logistic regression and classification metrics;
- PCA and dimensionality reduction;
- decision trees, random forests, and XGBoost.

Now we move to another family of methods:

- K-nearest neighbors;
- support vector machines;
- clustering.

These methods introduce three important ideas:

1. **Distance** between observations.
2. **Decision boundaries** in feature space.
3. **Unsupervised structure discovery**.

---

# 2. Dataset: Wisconsin Diagnostic Breast Cancer

We will continue using the Wisconsin Diagnostic Breast Cancer dataset.

Target:

- `M`: malignant;
- `B`: benign.

Predictors:

- 30 quantitative features extracted from digitized images of cell nuclei.

## Load data

```{r}
wdbc_raw <- read_csv("breast-cancer.csv")

names(wdbc_raw) <- names(wdbc_raw) %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("-", "_") %>%
  str_replace_all("\\\\.", "_") %>%
  str_to_lower()

wdbc <- wdbc_raw %>%
  mutate(
    diagnosis = factor(diagnosis, levels = c("M", "B"))
  ) %>%
  select(-id)

glimpse(wdbc)
```



## Class balance

```{r}
wdbc %>%
  count(diagnosis) %>%
  mutate(percent = n / sum(n) * 100)
```

```{r}
ggplot(wdbc, aes(x = diagnosis, fill = diagnosis)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Class balance",
    x = "Diagnosis",
    y = "Count"
  )
```

---

# 3. Train/test split

We keep a separate test set for honest evaluation.

```{r}
set.seed(22)

train_data <- wdbc %>%
  group_by(diagnosis) %>%
  slice_sample(prop = 0.75) %>%
  ungroup()

test_data <- wdbc %>%
  anti_join(train_data, by = names(wdbc))

train_data %>% count(diagnosis)
test_data %>% count(diagnosis)
```

---

# 4. Why scaling matters

KNN and SVM are distance-based or geometry-based methods.

If one feature has a much larger numeric scale than another, it can dominate the model.

Example:

- `area_mean` may be hundreds or thousands;
- `smoothness_mean` may be around 0.1.

Without scaling, the model may mostly pay attention to `area_mean`.

```{r}
wdbc %>%
  select(-diagnosis) %>%
  summarise(across(everything(), sd)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "sd"
  ) %>%
  arrange(desc(sd)) %>%
  head(10)
```


## Scale using train data only

```{r}
x_train <- train_data %>% select(-diagnosis)
x_test <- test_data %>% select(-diagnosis)

y_train <- train_data$diagnosis
y_test <- test_data$diagnosis

train_means <- x_train %>%
  summarise(across(everything(), mean))

train_sds <- x_train %>%
  summarise(across(everything(), sd))

x_train_scaled <- scale(
  x_train,
  center = as.numeric(train_means),
  scale = as.numeric(train_sds)
)

x_test_scaled <- scale(
  x_test,
  center = as.numeric(train_means),
  scale = as.numeric(train_sds)
)
```

# pandas equivalent:
# scaler = StandardScaler()
# X_train_scaled = scaler.fit_transform(X_train)
# X_test_scaled = scaler.transform(X_test)

---

# 5. Helper function for evaluation

```{r}
evaluate_classification <- function(truth, predicted_class, prob_malignant = NULL) {
  eval_data <- tibble(
    truth = factor(truth, levels = c("M", "B")),
    estimate = factor(predicted_class, levels = c("M", "B"))
  )

  metrics <- metric_set(accuracy, precision, recall, spec, f_meas)(
    eval_data,
    truth = truth,
    estimate = estimate
  ) %>%
    select(.metric, .estimate) %>%
    pivot_wider(names_from = .metric, values_from = .estimate)

  if (!is.null(prob_malignant)) {
    auc_value <- roc(
      response = factor(truth, levels = c("B", "M")),
      predictor = prob_malignant,
      levels = c("B", "M")
    ) %>%
      auc() %>%
      as.numeric()

    metrics <- metrics %>% mutate(auc = auc_value)
  }

  metrics
}
```



# 6. K-nearest neighbors

KNN is one of the simplest classification algorithms.

For a new sample:

1. calculate distance to training samples;
2. find the `k` nearest neighbors;
3. use majority vote among those neighbors;
4. assign the most common class.

Example:

```text
New sample
  nearest neighbors:
    M, M, B, M, B
  majority = M
  prediction = malignant
```

## Why `k` matters

Small `k`:

- very flexible;
- sensitive to noise;
- can overfit.

Large `k`:

- smoother decision boundary;
- less sensitive to noise;
- can underfit.



## KNN with k = 5

```{r}
set.seed(22)

knn_pred_k5 <- knn(
  train = x_train_scaled,
  test = x_test_scaled,
  cl = y_train,
  k = 5,
  prob = TRUE
)

knn_pred_k5 <- factor(knn_pred_k5, levels = c("M", "B"))

evaluate_classification(
  truth = y_test,
  predicted_class = knn_pred_k5
)
```

Note:

The base `class::knn()` function gives class predictions but does not provide convenient class probabilities for ROC AUC.

For simplicity, we evaluate KNN using threshold-based class metrics.


## Confusion matrix

```{r}
table(
  Truth = y_test,
  Prediction = knn_pred_k5
)
```



```{r}
k_values <- c(1, 3, 5, 7, 11, 15, 21)

knn_results <- map_dfr(
  k_values,
  function(k_value) {
    pred <- knn(
      train = x_train_scaled,
      test = x_test_scaled,
      cl = y_train,
      k = k_value
    )

    evaluate_classification(
      truth = y_test,
      predicted_class = factor(pred, levels = c("M", "B"))
    ) %>%
      mutate(k = k_value)
  }
)

knn_results
```

```{r}
knn_results %>%
  pivot_longer(
    cols = c(accuracy, precision, recall, spec, f_meas),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = k, y = value, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Effect of k on KNN performance",
    x = "Number of neighbors k",
    y = "Metric value",
    color = "Metric"
  )
```


# 7. Visualizing KNN idea in two dimensions

High-dimensional KNN is hard to visualize.  
For illustration, we use only two features.

```{r}
ggplot(train_data, aes(x = radius_mean, y = texture_mean, color = diagnosis)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Two-dimensional view of WDBC training data",
    x = "Mean radius",
    y = "Mean texture",
    color = "Diagnosis"
  )
```

KNN uses nearby points in feature space.  
In real use, it considers all scaled features, not only two.

---

# 8. Support Vector Machines

Support Vector Machines try to find a decision boundary that separates classes.

In two dimensions, this boundary is a line.

The goal is not just to separate classes, but to maximize the **margin**.

The margin is the distance between the decision boundary and the closest samples.

These closest samples are called **support vectors**.

## Linear SVM intuition

```text
Class A       |       Class B
   o          |          x
      o       |       x
        o     |    x
--------------|----------------  decision boundary
```

A wider margin usually means better generalization.

---

## Fit linear SVM

```{r}
svm_train_df <- as_tibble(x_train_scaled) %>%
  mutate(diagnosis = y_train)

svm_test_df <- as_tibble(x_test_scaled) %>%
  mutate(diagnosis = y_test)

linear_svm <- svm(
  diagnosis ~ .,
  data = svm_train_df,
  kernel = "linear",
  probability = TRUE
)

linear_svm
```

# pandas equivalent:
# svm_model = SVC(kernel="linear", probability=True)
# svm_model.fit(X_train_scaled, y_train)

---

## Predict with linear SVM

```{r}
linear_svm_pred <- predict(
  linear_svm,
  newdata = svm_test_df,
  probability = TRUE
)

linear_svm_prob <- attr(linear_svm_pred, "probabilities")[, "M"]

linear_svm_results <- evaluate_classification(
  truth = y_test,
  predicted_class = factor(linear_svm_pred, levels = c("M", "B")),
  prob_malignant = linear_svm_prob
) %>%
  mutate(model = "Linear SVM")

linear_svm_results
```

---

# 9. Kernel trick and radial SVM

A linear SVM can only draw a linear decision boundary.

But many datasets are not linearly separable.

The **kernel trick** allows SVMs to create nonlinear decision boundaries.

Common kernels:

- linear;
- polynomial;
- radial basis function.

The radial kernel is often a strong default for nonlinear data.

---

## Fit radial SVM

```{r}
radial_svm <- svm(
  diagnosis ~ .,
  data = svm_train_df,
  kernel = "radial",
  probability = TRUE
)

radial_svm
```

---

## Predict with radial SVM

```{r}
radial_svm_pred <- predict(
  radial_svm,
  newdata = svm_test_df,
  probability = TRUE
)

radial_svm_prob <- attr(radial_svm_pred, "probabilities")[, "M"]

radial_svm_results <- evaluate_classification(
  truth = y_test,
  predicted_class = factor(radial_svm_pred, levels = c("M", "B")),
  prob_malignant = radial_svm_prob
) %>%
  mutate(model = "Radial SVM")

radial_svm_results
```

---

# 10. SVM hyperparameters

Important SVM hyperparameters:

## Cost

`cost` controls how strongly the model penalizes misclassification.

Low cost:

- wider margin;
- more tolerance for errors;
- smoother boundary.

High cost:

- tries harder to classify training samples correctly;
- narrower margin;
- can overfit.

## Gamma

For radial SVM, `gamma` controls how local the decision boundary is.

Low gamma:

- smoother decision boundary.

High gamma:

- more flexible boundary;
- can overfit.

---

## Simple cost comparison

```{r}
cost_values <- c(0.1, 1, 10)

svm_cost_results <- map_dfr(
  cost_values,
  function(cost_value) {
    model <- svm(
      diagnosis ~ .,
      data = svm_train_df,
      kernel = "radial",
      cost = cost_value,
      probability = TRUE
    )

    pred <- predict(
      model,
      newdata = svm_test_df,
      probability = TRUE
    )

    prob <- attr(pred, "probabilities")[, "M"]

    evaluate_classification(
      truth = y_test,
      predicted_class = factor(pred, levels = c("M", "B")),
      prob_malignant = prob
    ) %>%
      mutate(cost = cost_value)
  }
)

svm_cost_results
```

---

# 11. Compare supervised models

```{r}
best_knn <- knn_results %>%
  arrange(desc(f_meas)) %>%
  slice(1) %>%
  mutate(model = paste0("KNN k=", k)) %>%
  select(model, accuracy, precision, recall, spec, f_meas)

supervised_results <- bind_rows(
  best_knn,
  linear_svm_results %>% select(model, accuracy, precision, recall, spec, f_meas, auc),
  radial_svm_results %>% select(model, accuracy, precision, recall, spec, f_meas, auc)
)

supervised_results
```

Discussion:

- Which model has the highest recall?
- Which model has the highest precision?
- Which model would you prefer for screening?
- Which model is easiest to explain?


---

# 12. From supervised to unsupervised learning

KNN and SVM are supervised:

- they use known labels during training.

K-means clustering is unsupervised:

- it does not use labels;
- it tries to discover groups based only on feature similarity.

This is useful when we do not know the true labels.

Examples:

- patient subgroups;
- cell populations;
- image phenotype clusters;
- molecular descriptor groups;
- experimental batch structure.

---

# 13. K-means clustering

K-means tries to divide data into `k` clusters.

Algorithm:

1. choose number of clusters `k`;
2. initialize cluster centers;
3. assign each point to nearest center;
4. update centers;
5. repeat until stable.

Important:

- k-means uses distance;
- scaling is essential;
- we must choose `k`;
- clusters are not automatically biological classes.

---

## K-means with k = 2

Since the WDBC dataset has two known classes, we can try `k = 2`.

However, k-means does not know the labels.

```{r}
set.seed(42)

kmeans_2 <- kmeans(
  x_train_scaled,
  centers = 2,
  nstart = 25
)

kmeans_2
```

---

## Compare clusters with true labels

```{r}
cluster_df <- tibble(
  diagnosis = y_train,
  cluster = factor(kmeans_2$cluster)
)

table(
  Diagnosis = cluster_df$diagnosis,
  Cluster = cluster_df$cluster
)
```

Important:

Cluster number `1` or `2` has no inherent meaning.  
We compare clusters to known labels only after clustering.

---

# 14. Visualize clusters with PCA

Because k-means was run in 30-dimensional space, we need a 2D projection for visualization.

We will use PCA only for plotting.

```{r}
pca_model <- prcomp(
  x_train_scaled,
  center = FALSE,
  scale. = FALSE
)

pca_scores <- as_tibble(pca_model$x[, 1:2]) %>%
  mutate(
    diagnosis = y_train,
    cluster = factor(kmeans_2$cluster)
  )
```

---

## PCA plot colored by true diagnosis

```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2, color = diagnosis)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA projection colored by true diagnosis",
    x = "PC1",
    y = "PC2",
    color = "Diagnosis"
  )
```

---

## PCA plot colored by k-means cluster

```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA projection colored by k-means cluster",
    x = "PC1",
    y = "PC2",
    color = "Cluster"
  )
```

Question:

> Do the unsupervised clusters match the known diagnosis labels?

---

# 15. Choosing number of clusters

In real unsupervised learning, we usually do not know the true number of clusters.

One simple method is the elbow plot.

K-means returns total within-cluster sum of squares.

Lower is better, but it always decreases as `k` increases.

We look for an elbow point where improvement slows.

```{r}
set.seed(2)

kmeans_wss <- map_dfr(
  1:10,
  function(k_value) {
    model <- kmeans(
      x_train_scaled,
      centers = k_value,
      nstart = 25
    )

    tibble(
      k = k_value,
      total_withinss = model$tot.withinss
    )
  }
)

kmeans_wss
```

```{r}
ggplot(kmeans_wss, aes(x = k, y = total_withinss)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Elbow plot for k-means",
    x = "Number of clusters k",
    y = "Total within-cluster sum of squares"
  )
```

---

# 16. Strengths and weaknesses

## KNN

Strengths:

- simple;
- no explicit training;
- flexible boundaries;
- useful baseline.

Weaknesses:

- slow prediction for large datasets;
- sensitive to scaling;
- sensitive to irrelevant features;
- choice of `k` matters;
- less interpretable in high dimensions.

## SVM

Strengths:

- strong classifier;
- works well in high-dimensional spaces;
- kernels allow nonlinear boundaries;
- often effective on small-to-medium datasets.

Weaknesses:

- scaling required;
- hyperparameters matter;
- probability estimates can be less direct;
- harder to interpret than logistic regression or trees.

## K-means

Strengths:

- simple;
- fast;
- useful for exploratory analysis;
- good for roughly spherical clusters.

Weaknesses:

- must choose `k`;
- sensitive to scaling;
- sensitive to initialization;
- assumes clusters are compact and roughly spherical;
- clusters may not match biological labels.
