---
title: "Practical: PCA and Dimensionality Reduction"
subtitle: "Wisconsin Diagnostic Breast Cancer dataset, feature correlation, PCA, t-SNE/UMAP, and logistic regression"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
---


```{r}
library(tidyverse)
library(broom)
library(yardstick)
library(pROC)
```


# pandas equivalent:
# import pandas as pd
# import numpy as np
# from sklearn.datasets import load_breast_cancer
# from sklearn.preprocessing import StandardScaler
# from sklearn.decomposition import PCA
# from sklearn.linear_model import LogisticRegression
# from sklearn.metrics import roc_auc_score, accuracy_score, precision_score, recall_score, f1_score
# from sklearn.model_selection import train_test_split

---

# 1. Why dimensionality reduction?

In the previous practical, we used logistic regression for binary classification.

That works well when we have a small number of predictors.

But real biomedical datasets often contain many variables:

- gene expression values;
- protein markers;
- microscopy-derived features;
- radiomics features;
- chemical descriptors;
- image embeddings;
- omics measurements.

Problems with many features:

1. Visualization becomes impossible.
2. Many variables are correlated.
3. Models become harder to interpret.
4. Noise can accumulate.
5. Training can become slower.
6. Overfitting becomes more likely.
7. Some methods perform poorly in high dimensions.

Dimensionality reduction tries to replace many original features with fewer new features while preserving important information.


# 2. Dataset: Wisconsin Diagnostic Breast Cancer

We will use the Wisconsin Diagnostic Breast Cancer dataset.

Each row describes a breast mass sample.

The target variable is:

- `M`: malignant;
- `B`: benign.

The predictors are 30 quantitative features computed from digitized images of cell nuclei.


## Load data

```{r}
wdbc_raw <- read_csv("Class 4/breast-cancer.csv")

glimpse(wdbc_raw)
```


## Clean column names

```{r}
names(wdbc_raw) <- names(wdbc_raw) %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("-", "_") %>%
  str_replace_all("\\.", "_") %>%
  str_to_lower()

wdbc <- wdbc_raw

glimpse(wdbc)
```

---

## Identify target and predictors

```{r}
wdbc <- wdbc %>%
  mutate(
    diagnosis = factor(diagnosis, levels = c("M", "B"))
  )

wdbc %>%
  count(diagnosis)
```

Important:

We set the first level to `M`, so in later classification metrics, malignant is treated as the positive class.

---

## Remove ID column

```{r}
wdbc_model_data <- wdbc %>%
  select(-id)

glimpse(wdbc_model_data)
```

---

# 3. Initial exploration

## Class balance

```{r}
wdbc_model_data %>%
  count(diagnosis) %>%
  mutate(percent = n / sum(n) * 100)
```

```{r}
ggplot(wdbc_model_data, aes(x = diagnosis, fill = diagnosis)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Class balance",
    x = "Diagnosis",
    y = "Count"
  )
```

---

## Number of features

```{r}
n_features <- wdbc_model_data %>%
  select(-diagnosis) %>%
  ncol()

n_features
```

Question:

> Can we easily visualize 30 features at once?

No. We can only easily plot 2 or 3 dimensions at a time.

---

## Example two-feature plot

```{r}
ggplot(
  wdbc_model_data,
  aes(x = radius_mean, y = texture_mean, color = diagnosis)
) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Only two features from a 30-feature dataset",
    x = "Mean radius",
    y = "Mean texture",
    color = "Diagnosis"
  )
```

This plot ignores 28 other variables.

Dimensionality reduction gives us a way to use all variables but visualize them in 2D.

---

# 4. Too many features in logistic regression

Let us start with a logistic regression model using all original features.

---

## Train/test split

We use a train/test split because we want to evaluate performance on unseen data.

If we test a model on the same data used for training, performance is often too optimistic.

The test set simulates new future samples.

```{r}
set.seed(42)

train_data <- wdbc_model_data %>%
  group_by(diagnosis) %>%
  slice_sample(prop = 0.75) %>%
  ungroup()

test_data <- wdbc_model_data %>%
  anti_join(train_data, by = names(wdbc_model_data))

train_data %>% count(diagnosis)
test_data %>% count(diagnosis)
```

# pandas equivalent:
# train, test = train_test_split(df, test_size=0.25, random_state=42, stratify=df["diagnosis"])

---

## Full logistic regression model

```{r}
full_logistic_model <- glm(
  diagnosis ~ .,
  data = train_data,
  family = binomial
)

summary(full_logistic_model)
```

Problems:

- many coefficients;
- difficult interpretation;
- many predictors are correlated;
- coefficients may be unstable;
- warning messages may occur because classes are very separable.

---

## Evaluate full model

```{r}
full_test_predictions <- test_data %>%
  mutate(
    prob_malignant = predict(
      full_logistic_model,
      newdata = test_data,
      type = "response"
    ),
    predicted_class = factor(
      if_else(prob_malignant >= 0.5, "M", "B"),
      levels = c("M", "B")
    )
  )

full_metric_data <- full_test_predictions %>%
  mutate(
    diagnosis = factor(diagnosis, levels = c("M", "B")),
    predicted_class = factor(predicted_class, levels = c("M", "B"))
  )

metric_set(accuracy, precision, recall, spec, f_meas)(
  full_metric_data,
  truth = diagnosis,
  estimate = predicted_class
)
```

```{r}
full_auc <- roc(
  response = full_test_predictions$diagnosis,
  predictor = full_test_predictions$prob_malignant,
  levels = c("B", "M")
) %>%
  auc()

full_auc
```

---

# 5. Correlation and redundancy

Many WDBC variables measure related properties.

For example:

- radius;
- perimeter;
- area;

are usually strongly correlated.

---

## Correlation matrix

```{r}
feature_data <- wdbc_model_data %>%
  select(-diagnosis)

cor_matrix <- cor(feature_data)

cor_matrix[1:5, 1:5]
```

---

## Correlation heatmap

```{r}
cor_long <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("feature_1") %>%
  pivot_longer(
    cols = -feature_1,
    names_to = "feature_2",
    values_to = "correlation"
  )

ggplot(cor_long, aes(x = feature_1, y = feature_2, fill = correlation)) +
  geom_tile() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6)
  ) +
  labs(
    title = "Correlation heatmap of original features",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  )
```

Interpretation:

- large blocks of high correlation suggest redundancy;
- PCA can compress correlated variables into fewer components.

---

# 6. PCA intuition

PCA finds new axes called **principal components**.

These components are:

- linear combinations of original variables;
- ordered by how much variance they explain;
- uncorrelated with each other.

PC1 explains the most variance.  
PC2 explains the next most variance, while being orthogonal to PC1.

---

## Synthetic example: two correlated variables

```{r}
set.seed(123)

pca_toy <- tibble(
  x = rnorm(200),
  y = x * 1.4 + rnorm(200, sd = 0.4)
)

ggplot(pca_toy, aes(x = x, y = y)) +
  geom_point(alpha = 0.7) +
  coord_equal() +
  theme_minimal() +
  labs(
    title = "Two correlated variables",
    subtitle = "Most variation lies along one diagonal direction"
  )
```

```{r}
toy_pca <- prcomp(pca_toy, center = TRUE, scale. = TRUE)

summary(toy_pca)
```

The first principal component captures most of the variance.

---

## Toy PCA scores

```{r}
toy_scores <- as_tibble(toy_pca$x)

ggplot(toy_scores, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.7) +
  coord_equal() +
  theme_minimal() +
  labs(
    title = "Same data in PCA coordinates",
    subtitle = "PC1 captures most variation; PC2 captures remaining variation"
  )
```

---

# 7. Scaling before PCA

Scaling is essential when variables have different units or ranges.

Without scaling, variables with large numerical ranges dominate PCA.

```{r}
feature_data %>%
  summarise(across(everything(), sd)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "sd"
  ) %>%
  arrange(desc(sd)) %>%
  head(10)
```

Because variables have different scales, we will standardize them.

---

# 8. PCA on WDBC data

Important workflow:

1. Split train/test.
2. Learn scaling parameters on training data.
3. Apply the same scaling to test data.
4. Fit PCA on training data.
5. Project test data using training PCA.

This avoids test data leakage.

---

## Prepare train and test matrices

```{r}
x_train <- train_data %>%
  select(-diagnosis)

x_test <- test_data %>%
  select(-diagnosis)

y_train <- train_data$diagnosis
y_test <- test_data$diagnosis
```

---

## Scale training data

```{r}
train_means <- x_train %>%
  summarise(across(everything(), mean))

train_sds <- x_train %>%
  summarise(across(everything(), sd))

x_train_scaled <- scale(
  x_train,
  center = as.numeric(train_means),
  scale = as.numeric(train_sds)
)

x_test_scaled <- scale(
  x_test,
  center = as.numeric(train_means),
  scale = as.numeric(train_sds)
)
```

# pandas equivalent:
# scaler = StandardScaler()
# X_train_scaled = scaler.fit_transform(X_train)
# X_test_scaled = scaler.transform(X_test)

---

## Run PCA on training data only

```{r}
pca_model <- prcomp(
  x_train_scaled,
  center = FALSE,
  scale. = FALSE
)

summary(pca_model)
```

We already scaled the matrix, so `center = FALSE` and `scale. = FALSE`.

---

# 9. Explained variance

Each principal component explains some fraction of total variance.

```{r}
explained_variance <- pca_model$sdev^2 / sum(pca_model$sdev^2)

explained_variance_df <- tibble(
  component = paste0("PC", seq_along(explained_variance)),
  component_number = seq_along(explained_variance),
  explained_variance = explained_variance,
  cumulative_variance = cumsum(explained_variance)
)

explained_variance_df %>%
  head(10)
```

---

## Scree plot

```{r}
ggplot(explained_variance_df, aes(x = component_number, y = explained_variance)) +
  geom_col(fill = "steelblue") +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Scree plot",
    subtitle = "Variance explained by each principal component",
    x = "Principal component",
    y = "Explained variance"
  )
```

---

## Cumulative variance plot

```{r}
ggplot(explained_variance_df, aes(x = component_number, y = cumulative_variance)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Cumulative explained variance",
    subtitle = "How many PCs are needed to explain 90% of variance?",
    x = "Number of principal components",
    y = "Cumulative variance"
  )
```

```{r}
n_pc_90 <- explained_variance_df %>%
  filter(cumulative_variance >= 0.90) %>%
  slice(1) %>%
  pull(component_number)

n_pc_90
```

---

# 10. PCA visualization

Now we project the training and test sets into PCA coordinates.

```{r}
train_pca_scores <- as_tibble(predict(pca_model, newdata = x_train_scaled)) %>%
  mutate(
    diagnosis = y_train,
    dataset = "train"
  )

test_pca_scores <- as_tibble(predict(pca_model, newdata = x_test_scaled)) %>%
  mutate(
    diagnosis = y_test,
    dataset = "test"
  )

all_pca_scores <- bind_rows(train_pca_scores, test_pca_scores)
```

---

## PC1 vs PC2

```{r}
ggplot(all_pca_scores, aes(x = PC1, y = PC2, color = diagnosis, shape = dataset)) +
  geom_point(alpha = 0.8, size = 2.5) +
  theme_minimal() +
  labs(
    title = "WDBC data projected onto first two principal components",
    subtitle = "PCA uses all original features to create PC1 and PC2",
    x = "PC1",
    y = "PC2",
    color = "Diagnosis",
    shape = "Dataset"
  )
```

This 2D plot uses information from all 30 original variables.

---

# 11. PCA loadings

PCA loadings tell us how strongly each original variable contributes to each principal component.

```{r}
loadings_df <- as.data.frame(pca_model$rotation) %>%
  rownames_to_column("feature") %>%
  as_tibble()

loadings_df %>%
  select(feature, PC1, PC2) %>%
  head()
```

---

## Top contributors to PC1

```{r}
loadings_df %>%
  select(feature, PC1) %>%
  mutate(abs_loading = abs(PC1)) %>%
  arrange(desc(abs_loading)) %>%
  slice(1:10)
```

```{r}
loadings_df %>%
  select(feature, PC1) %>%
  mutate(abs_loading = abs(PC1)) %>%
  arrange(desc(abs_loading)) %>%
  slice(1:10) %>%
  ggplot(aes(x = reorder(feature, abs_loading), y = abs_loading)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top contributors to PC1",
    x = "Feature",
    y = "Absolute loading"
  )
```

---

## Top contributors to PC2

```{r}
loadings_df %>%
  select(feature, PC2) %>%
  mutate(abs_loading = abs(PC2)) %>%
  arrange(desc(abs_loading)) %>%
  slice(1:10) %>%
  ggplot(aes(x = reorder(feature, abs_loading), y = abs_loading)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top contributors to PC2",
    x = "Feature",
    y = "Absolute loading"
  )
```

Discussion:

- PC1 often captures overall tumor/nucleus size and severity-related variation.
- PC2 may capture a different contrast among shape/texture-related measurements.
- PCA components are mathematically defined, so interpretation requires caution.

---

# 12. Logistic regression with PCA features

Now we compare:

1. logistic regression using all 30 original features;
2. logistic regression using first 2 PCs;
3. logistic regression using first 5 PCs;
4. logistic regression using first 10 PCs;
5. logistic regression using enough PCs for 90% variance.

---

## Helper function for model evaluation

```{r}
evaluate_binary_model <- function(data, truth_col, prob_col, threshold = 0.5) {
  truth <- data[[truth_col]]
  prob <- data[[prob_col]]

  pred <- factor(
    if_else(prob >= threshold, "M", "B"),
    levels = c("M", "B")
  )

  eval_data <- tibble(
    truth = factor(truth, levels = c("M", "B")),
    estimate = pred,
    prob = prob
  )

  metrics <- metric_set(accuracy, precision, recall, spec, f_meas)(
    eval_data,
    truth = truth,
    estimate = estimate
  )

  auc_value <- roc(
    response = eval_data$truth,
    predictor = eval_data$prob,
    levels = c("B", "M")
  ) %>%
    auc() %>%
    as.numeric()

  metrics %>%
    select(.metric, .estimate) %>%
    pivot_wider(names_from = .metric, values_from = .estimate) %>%
    mutate(auc = auc_value)
}
```

---

## Prepare PCA datasets

```{r}
make_pca_dataset <- function(scores, y, n_pc) {
  scores %>%
    select(all_of(paste0("PC", 1:n_pc))) %>%
    mutate(diagnosis = y)
}

train_pca_2 <- make_pca_dataset(train_pca_scores, y_train, 2)
test_pca_2 <- make_pca_dataset(test_pca_scores, y_test, 2)

train_pca_5 <- make_pca_dataset(train_pca_scores, y_train, 5)
test_pca_5 <- make_pca_dataset(test_pca_scores, y_test, 5)

train_pca_10 <- make_pca_dataset(train_pca_scores, y_train, 10)
test_pca_10 <- make_pca_dataset(test_pca_scores, y_test, 10)

train_pca_90 <- make_pca_dataset(train_pca_scores, y_train, n_pc_90)
test_pca_90 <- make_pca_dataset(test_pca_scores, y_test, n_pc_90)
```

---

## Fit PCA logistic models

```{r}
model_pca_2 <- glm(diagnosis ~ ., data = train_pca_2, family = binomial)
model_pca_5 <- glm(diagnosis ~ ., data = train_pca_5, family = binomial)
model_pca_10 <- glm(diagnosis ~ ., data = train_pca_10, family = binomial)
model_pca_90 <- glm(diagnosis ~ ., data = train_pca_90, family = binomial)
```

---

## Predict test probabilities

```{r}
pca_predictions_2 <- test_pca_2 %>%
  mutate(prob_malignant = predict(model_pca_2, newdata = test_pca_2, type = "response"))

pca_predictions_5 <- test_pca_5 %>%
  mutate(prob_malignant = predict(model_pca_5, newdata = test_pca_5, type = "response"))

pca_predictions_10 <- test_pca_10 %>%
  mutate(prob_malignant = predict(model_pca_10, newdata = test_pca_10, type = "response"))

pca_predictions_90 <- test_pca_90 %>%
  mutate(prob_malignant = predict(model_pca_90, newdata = test_pca_90, type = "response"))
```

---

## Compare models

```{r}
model_results <- bind_rows(
  evaluate_binary_model(full_test_predictions, "diagnosis", "prob_malignant") %>%
    mutate(model = "Full original features", n_features = 30),

  evaluate_binary_model(pca_predictions_2, "diagnosis", "prob_malignant") %>%
    mutate(model = "PCA 2 components", n_features = 2),

  evaluate_binary_model(pca_predictions_5, "diagnosis", "prob_malignant") %>%
    mutate(model = "PCA 5 components", n_features = 5),

  evaluate_binary_model(pca_predictions_10, "diagnosis", "prob_malignant") %>%
    mutate(model = "PCA 10 components", n_features = 10),

  evaluate_binary_model(pca_predictions_90, "diagnosis", "prob_malignant") %>%
    mutate(model = "PCA 90% variance", n_features = n_pc_90)
) %>%
  select(model, n_features, accuracy, precision, recall, spec, f_meas, auc) %>%
  arrange(desc(auc))

model_results
```

---

## Visual comparison

```{r}
model_results %>%
  select(model, accuracy, precision, recall, spec, f_meas, auc) %>%
  pivot_longer(
    cols = -model,
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = reorder(model, value), y = value, fill = metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Original features vs PCA features",
    x = "Model",
    y = "Metric value",
    fill = "Metric"
  )
```

Discussion:

- Does PCA reduce dimensionality without major performance loss?
- How many PCs are enough?
- Is the full model easier or harder to interpret?
- Is PCA better for prediction, visualization, or both?

---

# 13. Training time comparison

For this small dataset, timing differences may be small, but the principle matters for larger datasets.

```{r}
time_model_fit <- function(data) {
  start <- Sys.time()
  model <- glm(diagnosis ~ ., data = data, family = binomial)
  end <- Sys.time()
  as.numeric(end - start, units = "secs")
}

timing_results <- tibble(
  model = c("Full original features", "PCA 2", "PCA 5", "PCA 10", "PCA 90%"),
  training_time_seconds = c(
    time_model_fit(train_data),
    time_model_fit(train_pca_2),
    time_model_fit(train_pca_5),
    time_model_fit(train_pca_10),
    time_model_fit(train_pca_90)
  )
)

timing_results
```

---

# 14. Beyond PCA

## PCA

Type:

- unsupervised;
- linear.

Goal:

- preserve variance.

Strengths:

- fast;
- deterministic;
- interpretable through loadings;
- useful before modeling;
- useful for compression.

Limitations:

- only captures linear structure;
- components may be hard to interpret;
- high variance does not always mean high class relevance.

---

## t-SNE

Type:

- unsupervised;
- nonlinear.

Goal:

- preserve local neighborhoods.

Good for:

- visualization;
- clusters;
- single-cell data;
- image embeddings.

Limitations:

- not usually used as predictive model input;
- axes are not directly interpretable;
- distances between distant clusters can be misleading;
- results depend on parameters and random seed.

---

## UMAP

Type:

- unsupervised;
- nonlinear.

Goal:

- preserve local structure and some global structure.

Good for:

- visualization;
- high-dimensional biological datasets;
- embeddings;
- large datasets.

Limitations:

- parameters affect results;
- axes are not directly interpretable;
- not as simple to explain as PCA.

---

# 15. Optional: t-SNE visualization

Run this section only if `Rtsne` is installed.

```{r}
# library(Rtsne)
#
# tsne_result <- Rtsne(
#   as.matrix(x_train_scaled),
#   dims = 2,
#   perplexity = 30,
#   verbose = FALSE
# )
#
# tsne_df <- as_tibble(tsne_result$Y) %>%
#   rename(tSNE1 = V1, tSNE2 = V2) %>%
#   mutate(diagnosis = y_train)
#
# ggplot(tsne_df, aes(x = tSNE1, y = tSNE2, color = diagnosis)) +
#   geom_point(alpha = 0.8) +
#   theme_minimal() +
#   labs(
#     title = "t-SNE visualization of WDBC training data",
#     x = "t-SNE 1",
#     y = "t-SNE 2"
#   )
```




```{r}
# library(umap)
# 
#
# umap_result <- umap(as.matrix(x_train_scaled))
#
# umap_df <- as_tibble(umap_result$layout) %>%
#   rename(UMAP1 = V1, UMAP2 = V2) %>%
#   mutate(diagnosis = y_train)
#
# ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = diagnosis)) +
#   geom_point(alpha = 0.8) +
#   theme_minimal() +
#   labs(
#     title = "UMAP visualization of WDBC training data",
#     x = "UMAP 1",
#     y = "UMAP 2"
#   )
```

