---
title: "Practical: Linear Regression in R"
subtitle: "Model building, assumptions, transformations, selection, and polynomial regression"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
---


# Setup

```{r}
library(tidyverse)
library(broom)
library(modelr)
```

```{r}
# pandas equivalent:
# import pandas as pd
# import numpy as np
# import seaborn as sns
# import matplotlib.pyplot as plt
# import statsmodels.api as sm
# from sklearn.metrics import mean_squared_error
```


# Dataset

We will use the built-in `mpg` dataset from `ggplot2`.

Each row describes one car model.  
Our target variable will be:

- `hwy`: highway fuel efficiency in miles per gallon.

Possible predictors include:

- `displ`: engine displacement;
- `cyl`: number of cylinders;
- `drv`: drivetrain;
- `class`: vehicle class;
- `year`: model year;
- `cty`: city fuel efficiency.

```{r}
mpg
```



```{r}
# pandas equivalent:
# df.info()
# df.describe(include="all")
```


# 1. Exploratory Data Analysis

Before fitting a model, we need to understand the data.

```{r}
mpg %>%
  count(class, sort = TRUE)
```



```{r}
# pandas equivalent:
# df["class"].value_counts()
```


```{r}
mpg %>%
  summarise(
    n = n(),
    mean_hwy = mean(hwy),
    sd_hwy = sd(hwy),
    min_hwy = min(hwy),
    max_hwy = max(hwy)
  )
```



```{r}
# pandas equivalent:
# df["hwy"].agg(["count", "mean", "std", "min", "max"])
```

## Distribution of the target variable

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of highway MPG",
    x = "Highway MPG",
    y = "Count" )
```


## Relationship between displacement and highway MPG

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class), alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Engine displacement vs highway MPG",
    x = "Engine displacement",
    y = "Highway MPG")
```



## Boxplot by vehicle class

```{r}
ggplot(mpg, aes(x = class, y = hwy, fill = class)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip() +
  labs(
    title = "Highway MPG by vehicle class",
    x = "Vehicle class",
    y = "Highway MPG")
```



# 2. Simple Linear Regression

We start with one predictor:

```{r}
model_simple <- lm(hwy ~ displ, data = mpg)

summary(model_simple)
```



```{r}
# pandas equivalent:
# X = sm.add_constant(df["displ"])
# model = sm.OLS(df["hwy"], X).fit()
# print(model.summary())
```


## Interpretation

The model has the form:

$$
hwy = beta_0 + beta_1 * displ
$$

Where:

- intercept is expected `hwy` when `displ = 0`;
- slope tells how much `hwy` changes when `displ` increases by 1 unit.

```{r}
tidy(model_simple)
glance(model_simple)
```



```{r}
# pandas equivalent:
# model.params
# model.rsquared
# model.rsquared_adj
```


## Visualising the fitted regression line

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(
    title = "Simple linear regression",
    x = "Engine displacement",
    y = "Highway MPG")
```



# 3. Regression Assumptions

Linear regression is simple and powerful, but it works best when several assumptions are reasonably satisfied.

In practice, we do not expect perfect assumptions. We check whether violations are serious enough to make the model misleading.

The key assumptions are:

1. **Linearity**: the expected value of the target changes linearly with predictors.
2. **Normality of residuals**: residuals are approximately normally distributed.
3. **Homoscedasticity**: residual variance is approximately constant.
4. **Independence**: observations are independent.
5. **No severe multicollinearity**: predictors are not almost duplicates of each other.
6. **No extreme influential points**: individual observations should not dominate the model.

---

## Visual intuition: good vs problematic regression data

The next synthetic examples are not used for modelling.  
They are only visual illustrations.

```{r}
set.seed(2026)

assumption_examples <- tibble(
  x = rep(seq(1, 10, length.out = 80), 3),
  case = rep(c("Good linear pattern", "Nonlinear pattern", "Changing variance"), each = 80)
) %>%
  mutate(
    y = case_when(
      case == "Good linear pattern" ~ 3 + 2 * x + rnorm(n(), 0, 2),
      case == "Nonlinear pattern" ~ 3 + 1.5 * x + 0.8 * x^2 + rnorm(n(), 0, 5),
      case == "Changing variance" ~ 3 + 2 * x + rnorm(n(), 0, x * 0.8)
    )
  )

ggplot(assumption_examples, aes(x = x, y = y)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ case, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Linear regression assumption examples",
    subtitle = "The red line is a linear fit",
    x = "Predictor",
    y = "Target"
  )
```


## Assumption 1: Linearity

The relationship between predictors and the target should be approximately linear.

In our simple model:

```{r}
model_simple <- lm(hwy ~ displ, data = mpg)
```

we check whether the residuals show a systematic pattern.

```{r}
plot(model_simple, which = 1)
```

Interpretation:

- random scatter around zero is good;
- curved patterns suggest nonlinearity;
- strong patterns suggest the model is missing something.



## Assumption 2: Normality of residuals

Residuals should be approximately normally distributed.

This matters most for:

- p-values;
- confidence intervals;
- small datasets.

It is less critical for pure prediction in large datasets.

```{r}
plot(model_simple, which = 2)
```

A histogram of residuals:

```{r}
residual_data <- tibble(residuals = residuals(model_simple))

ggplot(residual_data, aes(x = residuals)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of residuals",
    x = "Residual",
    y = "Count")
```


## Assumption 3: Homoscedasticity

Homoscedasticity means approximately constant residual variance.

```{r}
plot(model_simple, which = 3)
```

Problem signs:

- funnel shape;
- increasing spread;
- decreasing spread.

A direct residual plot can also help:

```{r}
ggplot(model_simple, aes(x = .fitted, y = abs(.resid))) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE) +
  theme_minimal() +
  labs(
    title = "Absolute residuals vs fitted values",
    x = "Fitted values",
    y = "Absolute residuals")
```

If residuals become larger as fitted values increase, the model has non-constant variance.

Possible responses:

- transform the target variable;
- transform predictors;
- use weighted regression;
- use robust standard errors;
- use a different model.



## Assumption 4: Independence

Independence means observations should not be repeated measurements or strongly dependent on each other.

For this dataset, we assume each row is an independent car model.

Examples where independence may fail:

- repeated measurements from the same patient;
- time series;
- spatial measurements;
- multiple technical replicates from the same experiment;
- multiple rows from the same car model under slightly different settings.

If independence is violated, ordinary linear regression may underestimate uncertainty.

Possible alternatives:

- mixed-effects models;
- time-series models;
- clustered standard errors.



## Assumption 5: Multicollinearity

Multicollinearity means predictors are strongly correlated with each other.

For example, `cty` and `hwy` are very similar fuel-efficiency measurements, so using `cty` to predict `hwy` can give artificially strong results.

We can inspect correlations between numeric variables:

```{r}
mpg %>%
  select(where(is.numeric)) %>%
  cor()
```

# pandas equivalent:
# df.select_dtypes("number").corr()

A visual correlation heatmap:

```{r}
mpg %>%
  select(where(is.numeric)) %>%
  cor() %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(
    cols = -var1,
    names_to = "var2",
    values_to = "correlation"
  ) %>%
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  theme_minimal() +
  labs(
    title = "Correlation between numeric variables",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  )
```

High multicollinearity can cause:

- unstable coefficients;
- large standard errors;
- coefficients that change strongly when another variable is added;
- difficulty interpreting individual predictors.

---

## Assumption 6: Outliers and influential observations

Some points may strongly affect the regression line.

```{r}
plot(model_simple, which = 4)
```

Cook's distance helps identify influential observations.

Outliers should not be removed automatically.

First ask:

- Is it a data entry error?
- Is it a valid but unusual observation?
- Does the model depend too much on it?
- Should we fit models with and without it and compare?



# 4. Multimodal Data and Simpson's Paradox

Before moving to multiple regression, we should discuss a common problem in real datasets:

A relationship that appears in the pooled dataset can disappear, weaken, or even reverse when we split the data into meaningful groups.

This is often called **Simpson's paradox**.

It is especially important when data contain:

- several biological groups;
- different experimental batches;
- sex-specific effects;
- different species;
- different treatment conditions;
- different instruments or laboratories.

In regression, this means that one global regression line can be misleading.


## EMultimodal distribution

```{r}
set.seed(2026)

multimodal_data <- tibble(
  group = rep(c("Group A", "Group B"), each = 100),
  x = c(
    rnorm(100, mean = 3, sd = 0.8),
    rnorm(100, mean = 8, sd = 0.8)
  ),
  y = c(
    rnorm(100, mean = 12, sd = 1.5),
    rnorm(100, mean = 25, sd = 1.5)
  )
)

ggplot(multimodal_data, aes(x = x, fill = group)) +
  geom_histogram(alpha = 0.6, bins = 25, position = "identity") +
  theme_minimal() +
  labs(
    title = "Multimodal predictor distribution",
    subtitle = "The pooled distribution is actually a mixture of two groups",
    x = "Predictor x",
    y = "Count"
  )
```

If we ignore the groups, we may fit one global trend that mostly reflects group separation rather than a true within-group relationship.

```{r}
ggplot(multimodal_data, aes(x = x, y = y)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(
    title = "One global regression line",
    subtitle = "The line mostly captures differences between groups",
    x = "Predictor x",
    y = "Target y")
```

Now color by group:

```{r}
ggplot(multimodal_data, aes(x = x, y = y, color = group)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Group-wise regression lines",
    subtitle = "Within each group, the relationship can be different from the pooled trend",
    x = "Predictor x",
    y = "Target y")
```


## Why this matters for real data

In biological, chemical, and biomedical datasets, groups may represent:

- cell type;
- sex;
- species;
- organ;
- batch;
- lab;
- dose group;
- treatment protocol;
- disease status.

A pooled regression may be technically correct but scientifically misleading.

Before trusting a regression model, always ask:

1. Are there meaningful subgroups?
2. Does the target distribution look multimodal?
3. Does the predictor distribution look multimodal?
4. Do relationships change after coloring by group?
5. Should group be added as a predictor?
6. Should we test an interaction term?

##Simpson's paradox with `mpg`

In the `mpg` dataset, check whether the relationship between `displ` and `hwy` changes by vehicle `class`.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(
    title = "Pooled relationship between displacement and highway MPG",
    x = "Engine displacement",
    y = "Highway MPG"
  )
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Relationship between displacement and highway MPG by class",
    x = "Engine displacement",
    y = "Highway MPG"
  )
```

Fit and compare:

```{r}
mpg_pooled <- lm(hwy ~ displ, data = mpg)
mpg_group_adjusted <- lm(hwy ~ displ + class, data = mpg)
mpg_interaction <- lm(hwy ~ displ * class, data = mpg)

glance(mpg_pooled)
glance(mpg_group_adjusted)
glance(mpg_interaction)
```


# 5. Multiple Linear Regression

Now we use several predictors.

```{r}
model_multiple <- lm(
  hwy ~ displ + cyl + drv + class + year, data = mpg)

summary(model_multiple)
```



```{r}
# pandas equivalent:
# model = sm.OLS.from_formula("hwy ~ displ + cyl + C(drv) + C(class) + year", data=df).fit()
# print(model.summary())
```


R automatically converts categorical predictors into dummy variables.
A categorical variable with 3 levels becomes several comparison terms in the model.

```{r}
mpg %>%
  distinct(drv)
```

```{r}
tidy(model_multiple)
```


Important values:
- `r.squared`: explained variance;
- `adj.r.squared`: R² adjusted for number of predictors;
- `AIC`: model quality criterion;
- `BIC`: stricter model quality criterion;
- `p.value`: global model p-value.


# 7. Data Transformation

Sometimes the raw variables are not ideal for linear regression.

Transformations can help when:

- relationships are nonlinear;
- residuals are not normally distributed;
- variance changes with the fitted value;
- predictors are on very different scales.

---

## Log transformation

```{r}
mpg_transformed <- mpg %>%
  mutate(
    log_displ = log(displ))

model_log <- lm(hwy ~ log_displ, data = mpg_transformed)

summary(model_log)
```


```{r}
ggplot(mpg_transformed, aes(x = log_displ, y = hwy)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm") +
  theme_minimal()
```

---

## Scaling numeric predictors

Scaling does not usually change model predictions, but it makes coefficients easier to compare.

```{r}
mpg_scaled <- mpg %>%
  mutate(
    displ_scaled = as.numeric(scale(displ)),
    cyl_scaled = as.numeric(scale(cyl)),
    year_scaled = as.numeric(scale(year))
  )

model_scaled <- lm(
  hwy ~ displ_scaled + cyl_scaled + year_scaled,
  data = mpg_scaled
)

summary(model_scaled)
```


# 8. Forward and Backward Selection

Model selection helps choose a smaller model from a larger set of possible predictors.

Common goals:

- reduce unnecessary predictors;
- avoid overfitting;
- improve interpretability;
- improve prediction on new data.

Here we use `step()`, which performs AIC-based model selection.


## Full and null models

```{r}
full_model <- lm(
  hwy ~ displ + cyl + drv + class + year + fl + trans,
  data = mpg)

null_model <- lm(hwy ~ 1, data = mpg)
```


## Backward selection

Backward selection starts with the full model and removes predictors.

```{r}
backward_model <- step(
  full_model,
  direction = "backward",
  trace = FALSE)

summary(backward_model)
```


## Forward selection

Forward selection starts with the null model and adds predictors.

```{r}
forward_model <- step(
  null_model,
  scope = formula(full_model),
  direction = "forward",
  trace = FALSE)

summary(forward_model)
```

---

## Both-direction selection

```{r}
both_model <- step(
  null_model,
  scope = formula(full_model),
  direction = "both",
  trace = FALSE)

summary(both_model)
```


## Comparing selected models

```{r}
model_comparison <- tibble(
  model = c("full", "backward", "forward", "both"),
  adj_r_squared = c(
    glance(full_model)$adj.r.squared,
    glance(backward_model)$adj.r.squared,
    glance(forward_model)$adj.r.squared,
    glance(both_model)$adj.r.squared
  ),
  AIC = c(
    AIC(full_model),
    AIC(backward_model),
    AIC(forward_model),
    AIC(both_model)))

model_comparison %>%
  arrange(AIC)
```


# 9. Polynomial Regression

The relationship between engine size and highway MPG is not perfectly linear.

Polynomial regression lets us model curves while still using linear regression machinery.

---

## Quadratic model

```{r}
model_poly2 <- lm(
  hwy ~ poly(displ, 2),
  data = mpg)

summary(model_poly2)
```

```{r}
# pandas equivalent:
# df["displ2"] = df["displ"] ** 2
# model = sm.OLS.from_formula("hwy ~ displ + displ2", data=df).fit()
```


```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.7) +
  geom_smooth(
    method = "lm",
    formula = y ~ poly(x, 2),
    se = TRUE
  ) +
  theme_minimal() +
  labs(
    title = "Polynomial regression: quadratic fit",
    x = "Engine displacement",
    y = "Highway MPG")
```

---

## Comparing linear and polynomial models

```{r}
model_linear <- lm(hwy ~ displ, data = mpg)
model_poly3 <- lm(hwy ~ poly(displ, 3), data = mpg)

tibble(
  model = c("linear", "quadratic", "cubic"),
  adj_r_squared = c(
    glance(model_linear)$adj.r.squared,
    glance(model_poly2)$adj.r.squared,
    glance(model_poly3)$adj.r.squared
  ),
  AIC = c(
    AIC(model_linear),
    AIC(model_poly2),
    AIC(model_poly3)))
```

## Important warning

A more complex model can fit the training data better but perform worse on new data.

This is overfitting.

Polynomial regression should be evaluated using test data or cross-validation, not only visual fit.

---

# 10. Mini Challenge

Build the best linear regression model you can for predicting `hwy`.

Rules:

- use only variables from `mpg`;
- do not use `cty`, because it is too close to the target;
- try at least three models;
- compare them by adjusted R², AIC, and test RMSE;
- inspect residual plots for your final model.
